---
title: "Changes in Pan-Ethnicity and Racial Boundaries in the United States based on Interracial Marriage Patterns, 1980-2018"
author:
- affiliation: University of Oregon, Sociology
  name: Aaron Gullickson
keywords: keywords
#FYI: the thanks and abstract are easier to view and edit in Atom which does proper linewrapping
thanks: Thanks to people and stuff
abstract: This is a test abstract
output:
  pdf_document: 
    citation_package: natbib
    fig_caption: yes
    template: ./resources/aog-latex-ms.tex
    keep_tex: no
  word_document:
    reference_docx: ./resources/aog_word_style.docx
fontfamily: mathpazo
fontsize: 11pt
anonymous: false #anonymize, double-space, and ragged right
endnotes: false #make footnotes endnotes
endfloat: false #move table and figures to the end
bibliography: ../project.bib
biblio-style: ./resources/ajs.bst
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, 
                      warning = FALSE, eval = TRUE,
                      fig.height=3.5)
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
load(here("analysis","output","models.RData"))
load(here("analysis","output","alternates.RData"))
load(here("analysis","output","models_bendog.RData"))
#you need to load kableExtra in the doc
options(kableExtra.latex.load_packages = FALSE)
options(knitr.table.format = "latex")
library(kableExtra)
library(ggtext)
library(stringr)

my_palette <- c("#7b3294","#008837","#a6dba0")

#instead of loading markets, which takes forever, hardcoding sample
#sizes but be sure to adjust if this changes
#load(here("analysis","output","markets_census.RData"))
#load(here("analysis","output","markets_acs_full.RData"))
#census_n <- sum(markets_census[[1]]$choice)
#acs_n <- sum(markets_acs_full[[1]]$choice)
census_n <- 285523
acs_n <- 503348

```

```{r theme, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, include=FALSE}
theme_mine <- function() {
  theme_bw()+
    theme(axis.title = element_text(size=10), 
          axis.text = element_text(size=8),
          legend.text=element_text(size=7),
          legend.title=element_text(size=8),
          legend.key.size = unit(0.8,"line"),
          strip.text = element_text(size=9),
          legend.position = "bottom")
}
```

```{r create-coef-table}
#extract coefficient values from model summary as tibble
get_coefs <- function(x) {
  temp <- coef(x)[,c(1,2)]
  return(tibble(variable=rownames(temp), coef=temp[,1], se=temp[,2]))
}
#bind together the model summaries
bind_models <- function(models_summary) {
  temp <- lapply(models_summary, get_coefs) %>%
    bind_rows(.id="model") %>%
    filter(grepl("race_exog", variable))
  return(temp)
}

## Pentagon Models ##
coefs_pent <- list(census=models_census_pentagon,
                    acs_1980basis=models_acs1980_pentagon,
                    acs_full=models_acsfull_pentagon) %>%
  lapply(bind_models) %>%
  bind_rows(.id="data") %>%
  mutate(variable=convert_intermar_names(variable, "race_exog_pent"),
         model=factor(model,
                      levels=c("base","bendog","lendog","both"),
                      labels=c("Baseline model", "+ birthplace\nendogamy",
                               "+ language\nendogamy", "+ birthplace and\nlanguage endogamy")),
         data=factor(data,
                     levels=c("census","acs_1980basis","acs_full"),
                     labels=c("Census 1980", 
                              "ACS 2014-18, 1980 ethnic groups",
                              "ACS 2014-18, all ethnic groups"))) %>%
  order_variables("Census 1980", "+ both")

#create a dataset that can be used for the arrows in the graph
#need a reworked dataset for arrows
arrows_pent <- coefs_pent %>%
  filter(model=="+ birthplace and\nlanguage endogamy" & 
           variable!="South Asian ethnic exogamy" &
           data!="ACS 2014-18, all ethnic groups") %>%
  select(variable, coef, data) %>% 
  pivot_wider(names_from=data, values_from=coef) %>%
  rename(coef.census=`Census 1980`, 
         coef.acs=`ACS 2014-18, 1980 ethnic groups`)# %>%
  #filter(abs(exp(coef.census)-exp(coef.acs))>0.05)

## Extended Models ##
coefs_extended <- list(census=models_census_extended,
                       acs_1980basis=models_acs1980_extended,
                       acs_full=models_acsres_extended) %>%
  lapply(bind_models) %>%
  bind_rows(.id="data") %>%
  mutate(variable=convert_intermar_names(variable, "race_exog_extended"),
         model=factor(model,
                      levels=c("base","bendog","lendog","both"),
                      labels=c("Baseline", "+ birthplace endogamy",
                               "+ language endogamy", "+ both")),
         data=factor(data,
                     levels=c("census","acs_1980basis","acs_full"),
                     labels=c("Census 1980", 
                              "ACS 2014-18, 1980 ethnic groups",
                              "ACS 2014-18, expanded ethnic groups"))) %>%
  order_variables("Census 1980", "+ both")

#arrows
arrows_extended <- coefs_extended %>%
  filter(model=="+ both" & 
           variable!="South Asian ethnic exogamy" &
           data!="ACS 2014-18, expanded ethnic groups") %>%
  select(variable, coef, data) %>% 
  pivot_wider(names_from=data, values_from=coef) %>%
  rename(coef.census=`Census 1980`, 
         coef.acs=`ACS 2014-18, 1980 ethnic groups`) %>%
  filter(abs(exp(coef.census)-exp(coef.acs))>0.05)
```

```{r diversity}
temp <- alternates_census %>%
  filter(race!="AIAN" & race!="White" & race!="Black") %>%
  mutate(group=case_when(
    race %in% c("Chinese","Korean","Japanese","Vietnamese",
                     "Filipino","Cambodian","Hmong","Laotian",
                     "Thai","Burmese","Indonesian","Malaysian") ~ "E&SE Asian",
    race %in% c("Asian Indian","Bangladeshi",
                     "Pakistani","Sri Lankan") ~ "South Asian",
    race %in% c("Mexican","Puerto Rican","Cuban","Costa Rican",
                "Guatemalan","Honduran","Nicaraguan","Panamanian",
                "Salvadorian","Argentinian","Bolivian","Chilean",
                "Colombian","Ecuadorian","Paraguayan","Peruvian",
                "Uruguayan","Venezuelan","Dominican") ~ "Latino"),
    group=factor(group, levels=c("E&SE Asian","South Asian","Latino")))

diversity_ethnic <- temp %>%
  count(race, group, bpld) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980")

diversity_ethnic <- temp %>%
  count(race, group, lang) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980") %>%
  full_join(diversity_ethnic)

diversity_pan <- temp %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980")

diversity_pan <- temp %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980") %>%
  full_join(diversity_pan)

temp <- alternates_acs %>%
  filter(race!="AIAN" & race!="White" & race!="Black") %>%
  mutate(group=case_when(
    race %in% c("Chinese","Korean","Japanese","Vietnamese",
                     "Filipino","Cambodian","Hmong","Laotian",
                     "Thai","Burmese","Indonesian","Malaysian") ~ "E&SE Asian",
    race %in% c("Asian Indian","Bangladeshi",
                     "Pakistani","Sri Lankan") ~ "South Asian",
    race %in% c("Mexican","Puerto Rican","Cuban","Costa Rican",
                "Guatemalan","Honduran","Nicaraguan","Panamanian",
                "Salvadorian","Argentinian","Bolivian","Chilean",
                "Colombian","Ecuadorian","Paraguayan","Peruvian",
                "Uruguayan","Venezuelan","Dominican") ~ "Latino"),
    group=factor(group, levels=c("E&SE Asian","South Asian","Latino")))

temp2 <- temp %>%
  count(race, group, bpld) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18")

diversity_ethnic <- temp %>%
  count(race, group, lang) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18") %>%
  full_join(temp2) %>%
  bind_rows(diversity_ethnic) %>%
  select(race, group, data, diversity_bpl, diversity_lang)

temp2 <- temp %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\nall ethnic groups")

diversity_pan <- temp %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\nall ethnic groups") %>%
  full_join(temp2) %>%
  bind_rows(diversity_pan)

races1980 <- as.character(subset(diversity_ethnic, data=="Census 1980")$race)

temp2 <- temp %>%
  filter(race %in% races1980) %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups")

diversity_pan <- temp %>%
  filter(race %in% races1980) %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups") %>%
  full_join(temp2) %>%
  bind_rows(diversity_pan)  %>%
  mutate(type="pan-ethnic group diversity")

temp2 <- diversity_ethnic %>%
  group_by(group, data) %>%
  summarize(diversity_bpl=mean(diversity_bpl),
            diversity_lang=mean(diversity_lang)) %>%
  ungroup() %>%
  mutate(data=ifelse(data=="ACS 2014-18", "ACS 2014-18\nall ethnic groups",
                     data))

diversity_pan <- diversity_ethnic %>%
  filter(race %in% races1980 & data=="ACS 2014-18") %>%
  group_by(group, data) %>%
  summarize(diversity_bpl=mean(diversity_bpl),
            diversity_lang=mean(diversity_lang)) %>%
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups") %>%
  bind_rows(temp2)  %>%
  mutate(type="mean of within ethnic group diversity") %>%
  bind_rows(diversity_pan) %>%
  pivot_longer(cols=starts_with("diversity_"), names_to="variable",
               values_to="diversity") %>%
  mutate(variable=ifelse(variable=="diversity_bpl",
                     "Birthplace",
                     "Language"),
         diversity=ifelse(group=="South Asian" & 
                            data!="ACS 2014-18\nall ethnic groups",
                          NA, diversity),
         data=factor(data,
                     levels=c("Census 1980", 
                              "ACS 2014-18\n1980 ethnic groups",
                              "ACS 2014-18\nall ethnic groups")),
         group=factor(group, 
                      levels=c("E&SE Asian", "Latino","South Asian")))
```

# Introduction

Social scientists have long used patterns of intermarriage between groups to measure the underlying social distance and strength of boundaries separating groups. Because marriage is one of the most intimate kinds of social relationships, a widespread willingness to cross group boundaries is seen as a key signifier that the salience of those boundaries is weak [@gordon_assimilation_1964].

Typically, an increase in intermarriage across one boundary has been seen to indicate a breakdown of group boundaries, but more broadly it can also indicate a reshaping of existing boundaries. Widespread intermarriage among European ethnic groups in the mid-twentieth century contributed to the breakdown of salient ethnic divisions between these populations but also helped to re-consolidate a sense of collective whiteness among these groups [@lieberson_many_1988; @alba_ethnic_1990; @jacobsen_whiteness_1998]. Similarly, spurred by resurgent immigration to the US since the 1960s, scholars today have frequently been concerned with how patterns of intermarriage may contribute to our understanding of panethnic identity among Asian and Latino ethnic groups (CITATIONS?).[^1]

[^1]: I use the term "ethnic groups" here to distinguish primarily different nationality groups among the Asian and Latino population, such as Chinese, Korean, Mexican, and Colombian.

Panethnicity is defined by @okamoto_panethnicity_2014a as "the construction of a new categorical boundary through the consolidation of ethnic, tribal, religious, or national groups." More generally, we can see pan-ethnicity as a form of boundary expansion in which the salient boundary shifts from one level to another within a nested hierarchy [@wimmer_making_2008]. From this boundary perspective, ethnoracial categories are not fixed, stable, or primordial, but can shift over time in response to historical events. In US history, the categories of white, black, and American Indian have all become salient through the consolidation of smaller units of identification (CITE). Understanding how such processes are playing out for contemporary Asian and Latino ethnic groups informs us about how political and social boundaries might shift in the future.

Prior work on intermarriage among Asian and Latino ethnic groups has contributed greatly to our understanding of the phenomenon of panethnicity, but has also been hampered by two difficulties, one of which is data-based and the other conceptual. First, lack of information on the timing of marriage has made it difficult to address the "immigrants married abroad" (IMA) problem [@hwang_problem_1990]. Ideally, any analysis of intermarriage should exclude immigrants who were married elsewhere because those individuals would typically have had far less opportunity to marry outside of their ethnic/nationality group. Thus, inclusion of IMA will tend to deflate estimates of panethnic intermarriage. However, relatively few of the data sources uses to estimate intermarriage have included information on the timing of the current marriage, making it impossible to cleanly remove such cases. Scholars have adopted a wide variety of other strategies to minimize the effect of IMA on the anlysis, but it is difficult to know how much it might affect results.

Second, much of the work on intermarriage is framed around issues of assimilation and acculturation, and so panethnicity is often characterized in a similar way. Researchers focus on differences in pan-ethnic tendencies among the native-born and foreign-born. Panethnicity itself has also been characterized as a form of "selective acculturation" within the segemented assimilation framework [@bohra-mishra_intermarriage_2015]. I argue instead that we should think about these issues from a larger social boundary framework, and that panethnicity should be considered a distinct process of boundary expansion. In the current demographic environment, issues of accculturation affect the tendency toward panethnicity but should be thought of as distinct and separable processes, rather than integral components of panethnicity. Specifically, individuals tend to prefer to marry someone born (and raised) in the same place and who speaks the same language. Birthplace and language endogamy have a complicated relationship to the tendency toward panethnic marriage. Our goal should be to estimate that tendency toward panethnic marriage, accounting for these other tendencies rather than to collapse them all into an undifferentiated process of assimiliation/acculturation. Doing so allows us to see how much panethnic affinity there would be between panethnic groups in the absence of birthplace and language endogamy.

In this article, I address each of these difficulties to estimate the strength of pan-ethnicity among Asian and Latino ethnic groups at two points in time. I use Census 1980 and American Community Survey (ACS) data from 2014-18 to identify changes in pan-ethnic tendencies over time. Census 1980 was the last census to include information on the duration of marriage and duration of marriage has been included on the ACS since 2013. Thus, I am able to identify and exclude immigrants married abroad from analysis. Furthermore, I use a relatively novel counterfactual marriage model technique (CITE) that allows me to easily account for the role of birthplace and language endogamy in estimating the strength of panethnicity.

I find that the patterns of pan-ethnicity are very different for Asian and Latino ethnic groups. For Asians, panethnic affinity is very strong among East and Southeast Asian ethnic groups and among South Asian ethnic groups, but there is little evidence of pan-ethnicity between South Asian and East and Southeast Asian ethnic groups. Furthermore, this panethnic affinity has grown substantially over time. This panethnic affinity is reduced substantially in actual practice by significant language and birthplace differences across Asian ethnic groups. For Latinos, on the other hand, pan-ethnic affinity is much weaker and has changed little over time. Furthermore, it is exceeded in many cases by affinity for outmarriage to non-Latino racial groups. I argue that these different patterns have important implications for the future salience of these panethnic categories.

# Pan-Ethnicity in Marriage

* Panethnicity has been analyzed using a broad set of approaches. Much of the initial work focused on the development of panethnic organizations and movements. However, demographers in particular have been instrumental in thinking about intermarriage as a bellwether for actual affinities between ethnic groups, a part from political activism and organizational participation. Better measure of "on the ground" truth. 
* What drives panethnicity? It could be broader cultural affinity (such as language, in the case of Latinos), but most scholars argue it is primarily driven by processes of racialization and discrimination. Panethnic identification is often a result of struggles over classification between state agencies and ethnic leaders and entrepreneurs. (NOT SURE THIS GOES HERE BUT NEEDS TO GO SOMEWHERE) 
* Most important studies of panethnic intermarriage (Rosenfeld, Fu, Qian and Cobas, Qian, Blair, and Ruf) are over a decade old and use even older data. How do I fit Qian. Glick, and Baston in here?
  * Rosenfeld and Fu specifically compare patterns of panethnicity among Asians and Latinos and find a greater tendency for panethnicity among Asians than Latinos, which they both attribute to the role of racialization (CHECK). 
  * How likely is panethnic intermarriage relative to white/black outmarriage. Fu, and the two early Qian papers speak to this I believe. Fu clearly finds that ethnic exogamy is more likely for Asians than intermarriage with whites, and I think Qian, Blai, and Ruf find the same. Fu also finds ethnic exogamy for Latinos is not more likely than outmarriage to whites - not clear if Qian and Cobas find the same. I need to check to see if Rosenfeld explored this question.
  * How are South Asians handled? In many cases they are not included at all. Reasons to be suspect of the idea that Asian panethnicity will work the same for South Asians who are more commonly considered "ambiguous non-whites" than Asians in the US context. 
* Methodological difficulties of prior work:
  * Accounting for group size. group size matters (cite Blau) in who marries who, but is thought of as an issue of exposure. Most studies control for group size to get at underlying affinities. Studies that do not control for group size will seriously misunderstand the underlying propensities between groups because most results will be driven by group size differences (cite Min and Kim). Log-linear models have been the preferred method to address the group size issue.
  * Accounting for differential spatial patterns. Marriage markets are local and not national, so when adjusting for group size one needs to optimally adjust for group sizes in local marriage markets not just use national level distributions (cite Harris and Ono). This issue is particularly salient when discussing panethnicity among Latinos and Asians, because ethnic groups have different patterns of settlement that will restrict the opportunities for panethnic intermarriage (e.g. Cubans in Florida, PR in Northeast, and Mexicans in Southwest). Prior work has handled this issue in multiple ways. Some scholars have ignored the issue. Rosenfeld looked specifically at intermarriage within specific metropolitan areas. Fu controlled for group size within Census divisions.
  * The IMA issue and prevalence/incidence. Lack of information on the timing of marriage leads to two problems:
    * Scholars want to understand recent pattern of intermarriage but without timing of marriage information, it is impossible to limit analysis to only recently formed marriages. Scholars typically get around this by restricting the age of partners, which is less than ideal.
    * To understand intermarriage patterns in the US, you want to exclude IMA. However, without timing of marriage information, it is impossible to cleanly remove IMA from the analysis. Discuss workarounds by authors.  (probably shrink this discussion intro and put it here).
    
## The Role of Birthplace and Language Endogamy

* Intermarriage work often framed around issues of assimilation and acculturation. Panethnicity is often framed in a similar way. Work often divides sample by native and foreign-born population for example. 
* Problematic issue of assimilation to the "mainstream." What is mainstream? Segmented assimiliation tries to address this issue, but panethnicity does not fit well. Some have described it as a form of "selective acculturation" but this seems misplaced as the focus of that alternative is to engage in closure at the level of ethnic/nationality group, not panethnic group. 
* The boundary approach would suggest we think about panethnicity as a specific process of boundary expansion, and not simply as a function of assimilation and acculturation. 
* Nonetheless, the possibility for panethnicity is deeply affected by issues of assimilation and acculturation. Most notably, birthplace and language endogamy will affect the likelihood of panethnicity in complex ways. 
* In general, we would expect birthplace endogamy to affect Asian and Latino groups the same. Birthplace endogamy will lower the odds of panethnicity when you have a substantial foreign-born population. On the other hand, we would expect language endogamy to affect Asian and Latino groups differently. Language differences across Asian nationality groups will lower the odds of panethnicity, particularly when you have a substantial foreign-born population or lack of English language acquisition among the native-born. On the other hand, because Latinos share a common language, you would not expect language endogamy to lower the odds of panethnicity and it may even increase those odds.
* The actuality is considerably more complex than this however, because ethnic/nationality groups themselves are also diverse in terms of birthplace and language. For example (PROVIDE SOME EXAMPLES). These differences will also increae barriers to ethnic endogamy. Thus, the real issue is the extent to which language and birthplace diversity is how much greater diversity is within the panethnic group than within a specific ethnic group.
* I think introduce Figure 1 here and talk about it. 
* Ideally, to estimate the underlying affinity for panethnicity, we want a model that can separate out these barriers to intermarriage resulting from language and birthplace endogamy. Such a model would give us a counterfactual case of how much panethnicity we would expect in a situation in which all Latinos and Asians are fully assimilated/acculturated to the US (i.e. born in the US and speak English as their primary language). 

* In the analysis that follows I ... FILL THE GAP!

# Data and Methods

The data for this analysis are derived from the microdata sample of the 1980 US Census and the American Community Survey (ACS) for 2014-2018. The ACS is an annual 1-in-100 survey of the United States population, conducted by the United States Census Bureau. The 1980 US Census is the last Census to include information about marriage timing. Information about marriage timing was added to the ACS in 2008. To increase sample size for smaller ethnoracial categories, I pool five years of ACS data from 2014-2018. Both data sources were extracted from the Integrated Public Use Microdata Series (IPUMS) system [@ruggles_ipums_2020].

I use a counterfactual marriage model [@qian_marriage_2018; @gullickson_counterfactual_2021] for the analysis that follows. In both data sources, I restrict the analysis to all opposite-sex marriages formed in the previous five years that were a first marriage for both partners. The restriction to opposite-sex first marriages is necessary for comparison because the Census 1980 only recorded marriage timing relative to first marriage and does not include same-sex unions. For each marriage, I then construct a choice set of one real union and twenty fictional unions. Fictional unions are created by sampling alternate partners for one randomly determined spouse from a pool of potential partners. I then use a conditional logit model to predict how partner characteristics relate to the likelihood of observing the true union, as follows: $$P_{ij}=\frac{e^{\mathbf{x}_{ij}\beta}}{\sum_{k=1}^J e^{\mathbf{x}_{ik}\beta}}$$ where $P_{ij}$ is the probability that union $j$ within choice set $i$ is the actual union. $J$ is the total number of unions in the choice set (in this case, twenty-one). The vector $\mathbf{x}_{ij}$ defines the characteristics of the union and the $\beta$ vector provides estimated log-odds ratios indicating how the odds of an actual union change with $\mathbf{x}_{ij}$. The model is estimated as a fixed-effects logistic regression model with fixed effects for each choice set.

The set of potential alternate partners from which fictional spouses are sampled is made up of all unmarried adults as well as those who were married in the previous five years. Because of uncertainty regarding timing of migration and marriage, individuals who migrated to the US or across state lines in the last five years are excluded from analysis as both actual and alternate marriage partners. To better capture the supply of potential partners in local marriage markets, alternate partners are sampled from among individuals residing in the same marriage market. I define marriage markets by the metropolitan area of residence when this can be identified for a respondent, and otherwise by the state of residence. Metropolitan area is not directly identifiable in the data used here due to confidentiality concerns, but can be deduced for many metropolitan areas based on the finest geographic identifier of the respondent (county group in Census 1980 and Public Use Microdata Area in the ACS). IPUMS provides metro area identification for respondents in cases where the county group or PUMA was mostly contained within the metropolitan area. Figure 1 shows the identifiable metropolitan areas used in the analysis for both time periods. Although not all metropolitan areas are identified in the data, most major metropolitan areas are included in both time periods. Coverage differs somewhat by period, with more extensive coverage in the later time period.

```{r, fig.cap="Metropolitan statistical areas identifiable in the data by data source. Fill colors are drawn with partial transparency so that darker areas indicate overlapping areas in the two data sources."}
#knitr::include_graphics(here("analysis","output","msa_map.pdf"), 
#                        auto_pdf = TRUE)
```

Because this approach relies upon a random sample of alternate partners, results will vary each time this sampling procedure is performed. To account for the added uncertainty due to sampling, I preform the sampling procedure separately five times to produce five separate analytical datasets. I then run parallel models on all five datasets and pool $\beta$ estimates and standard errors using the same methods as those employed for multiple imputation [@rubin_multiple_1987].

This counterfactual marriage model approach has several advantages over the more well known log-linear approach to analyzing assortative mating. Most importantly, because the researcher can specify the parameters for sampling alternate partners, it is relatively easy to adjust for sub-national differences in group composition. Smaller groups in the population are more likely to be exogamous simply because there are fewer within-group potential partners available. Assortative mating researchers typically desire to factor out such "supply side" issues when estimating the likelihood of group exogamy. @harris_how_2005 have demonstrated that the uneven spatial distribution of racial groups across the US plays a significant role in restricting racial exogamy and log-linear models estimated at the national level will not account for this unevenness. This issue is likely to play an important role in this analysis, because Asian and Latino ethnic groups are not evenly distributed across the US. I account for this uneven spatial distribution by sampling alternate partners from within the same metropolitan area (or state, for respondents outside of an identifiable metropolitan area) such that potential partners more closely match the composition of the local marriage market for the focal partner.

The counterfactual model also has two other notable advantages over the log-linear model. First, it includes the unmarried population within the sample of alternate partners whereas the log-linear model only includes those who were married. Second, the linear model framework makes incorporation of a variety of quantitative and categorical control variables much easier.

## Variable Construction

I measure ethnoracial categories by a combination of the race and Hispanicity questions in both data sources. Table 1 shows the sample size for all ethnoracial groups included in the analysis. I classify non-Asian and non-Latino respondents as White, Black, or American Indian/Alaska Native (AIAN). Due to small sample size, I exclude Pacific Islanders, multriacial respondents, and non-Latino others from the analysis. The race question includes several Asian nationalities (e.g. Chinese, Korean, Filipino) as well as a write-in response for ethnic identities not captured by the existing categories. Similarly, the Hispanicity question includes major Latino ethnic groups (e.g. Mexican, Puerto Rican, Cuban) as well as a write-in option. Importantly, the Census 1980 data includes a far more limited set of identifiable ethnic groups for both Asians and Latinos than the ACS data. In Census 1980, I can only identify Mexicans, Puerto Ricans, and Cubans among Latinos and Chinese, Japanese, Korean, Filipino, Vietnamese, and Asian Indian respondents among Asians. To address this restriction, I estimate two different kinds of models for the ACS data. In the first model, I restrict the data only to those ethnic groups that were available in the Census 1980 data to allow more direct comparison over time. I also estimate models using the full ACS data, but these are not directly comparable to the Census 1980 results.

```{r sample-size-table}

temp <- c(table(alternates_acs$race)) %>% enframe()
tab_samp <- c(table(alternates_census$race)) %>% 
  enframe() %>%
  right_join(temp, by="name") %>%
  rename(ethnorace=name, census=value.x, acs=value.y) %>%
  mutate(race_big=case_when(
    ethnorace=="Chinese" ~ "East and Southeast Asian",
    ethnorace=="Japanese" ~ "East and Southeast Asian",
    ethnorace=="Korean" ~ "East and Southeast Asian",
    ethnorace=="Filipino" ~ "East and Southeast Asian",
    ethnorace=="Vietnamese" ~ "East and Southeast Asian",
    ethnorace=="Cambodian" ~ "East and Southeast Asian",
    ethnorace=="Hmong" ~ "East and Southeast Asian",
    ethnorace=="Laotian" ~ "East and Southeast Asian",
    ethnorace=="Thai" ~ "East and Southeast Asian",
    ethnorace=="Burmese" ~ "East and Southeast Asian",
    ethnorace=="Indonesian" ~ "East and Southeast Asian",
    ethnorace=="Malaysian" ~ "East and Southeast Asian",
    ethnorace=="Bhutanese" ~ "East and Southeast Asian",
    ethnorace=="Nepalese" ~ "East and Southeast Asian",
    ethnorace=="Asian Indian"  ~ "South Asian",
    ethnorace=="Pakistani" ~ "South Asian",
    ethnorace=="Bangladeshi" ~ "South Asian",
    ethnorace=="Sri Lankan" ~ "South Asian",
    ethnorace=="Mexican" ~ "Latino", 
    ethnorace=="Cuban" ~ "Latino", 
    ethnorace=="Puerto Rican" ~ "Latino", 
    ethnorace=="Guatemalan" ~ "Latino", 
    ethnorace=="Salvadorian" ~ "Latino", 
    ethnorace=="Dominican" ~ "Latino", 
    ethnorace=="Colombian" ~ "Latino", 
    ethnorace=="Costa Rican" ~ "Latino", 
    ethnorace=="Honduran" ~ "Latino", 
    ethnorace=="Nicaraguan" ~ "Latino", 
    ethnorace=="Panamanian" ~ "Latino", 
    ethnorace=="Argentinian" ~ "Latino", 
    ethnorace=="Bolivian" ~ "Latino", 
    ethnorace=="Chilean" ~ "Latino", 
    ethnorace=="Ecuadorian" ~ "Latino", 
    ethnorace=="Peruvian" ~ "Latino", 
    ethnorace=="Venezuelan" ~ "Latino", 
    ethnorace=="Paraguayan" ~ "Latino", 
    ethnorace=="Uruguayan"  ~ "Latino"
  ))

temp <- tab_samp %>% group_by(race_big) %>%
  summarize(census=sum(census, na.rm=TRUE), acs=sum(acs, na.rm=TRUE)) %>%
  na.omit() %>%
  rename(ethnorace=race_big)

tab_samp <- tab_samp %>% select(ethnorace, census, acs) %>%
  bind_rows(temp)

temp <- tibble(ethnorace="Marriages in the previous year",
               census=census_n, acs=acs_n)

tab_samp <- tab_samp %>% 
  bind_rows(temp)
                    
ord <- c("Marriages in the previous year",
         "White","Black","AIAN","Latino",
         "Mexican","Puerto Rican","Cuban","Salvadorian","Dominican","Guatemalan",
         "Colombian","Honduran","Peruvian","Ecuadorian","Nicaraguan","Argentinian",
         "Venezuelan","Panamanian","Chilean","Costa Rican","Bolivian",
         "Uruguayan","Paraguayan",
         "East and Southeast Asian",
         "Chinese","Filipino","Vietnamese","Korean","Japanese","Cambodian",
         "Hmong","Laotian","Thai","Burmese","Indonesian","Malaysian",
         "South Asian",
         "Asian Indian","Pakistani","Bangladeshi","Sri Lankan")

tab_samp <- as.data.frame(tab_samp)
rownames(tab_samp) <- tab_samp$ethnorace
tab_samp <- tab_samp[ord,]
rownames(tab_samp) <- NULL

options(knitr.kable.NA = '')

tab_samp %>%
  mutate(census=format(census, big.mark=","),
         census=ifelse(census=="       NA","",census),
         acs=format(acs, big.mark=",")) %>%
  kbl(caption = "Sample size of marriages and alternate partners by data source and ethnoracial category.", booktabs=T, longtable=F,
      linesep = NULL,
      col.names=c("Category","Census 1980", "ACS 2014-2018"),
      align = c("l","r","r")) %>%
  kable_styling(full_width = TRUE) %>%
  add_indent(c(6:24,26:37,39:42)) %>%
  pack_rows("Marriages", 1, 1) %>%
  pack_rows("Alternate Partners", 2, 42) %>%
  column_spec(1, width = "6cm")
```

I model interracial and panethnic marriage using a set of dummy variables where the reference category is a racially and ethnically endogamous union. Because the number of ethnoracial categories is large, modeling every possible combination of these categories for exogamous unions is not feasible, so I apply two limits on the parameterization. First, when analyzing intermarriage with Whites, Blacks, and American Indians/Alaska Natives, I group Asian and Latino ethnic groups into the larger pan-ethnic categories of Latino, East/Southeast Asian, and South Asian. For example, a union involving a black and Mexican partner and a union involving a black and Colombian partner would both be coded as a Black/Latino union. Second, for most of the models that follow, I model the likelihood of ethnic exogamy using a single dummy variable to indicate any form of ethnic exogamy within the three pan-ethnic categories of Latino, East/Southeast Asian, and South Asian. This allows me to estimate the average likelihood of ethnic exogamy within these groups, but does not capture particular affinities of disaffinities among ethnic groups within the pan-ethnic category. In some of the models that follow, I relax this assumption for the largest ethnic groups in order to estimate such affinities and disaffinities in the more recent ACS data.

Because I separate South Asian from East/Southeast Asian groups and only one South Asian ethnic group was identifiable in the Census 1980 data, I cannot estimate pan-ethnic parameters over time for South Asians. Thus, most of the analysis that follows will focus on pan-ethnicity among East/Southeast Asians and Latinos. However, I do estimate and report pan-ethnic parameters for South Asians for some models using ACS data.

I account for language and birthplace endogamy with simple dummy variables indicating whether the two potential partners share the same primary language or birthplace, respectively. Primary language is determined by what language the respondent reported speaking at home. Because this variable is measured after a marriage has occurred, it may somewhat overestimate language endogamy among respondents. Some partners who spoke different languages when they met may have shifted the language spoken at home after marriage.

Birthplace endogamy is complicated by the "1.5" generation -- individuals who were born outside of the United States but migrated as children and whose formative experiences are at least partially defined by acculturation within the US. Within the coding scheme used here, I consider three possibilities for such individuals organized by the assumed degree of acculturation and assimilation. First such individuals could be considered birthplace endogamous only with individuals from the same actual birthplace. Second, such individuals could be considered birthplace endogamous with either an individual from their actual birthplace or the a person born in the USA. Third, such individuals could be considered only birthplace endogamous with other individuals born in the USA.

To test the accuracy of these three scenarios, I fit models using each coding scheme to the Census 1980 and full ACS data. Following @rumbaut_ages_2004a, I also divide the 1.5 generation into a "1.75" generation (those who arrived in the US before the age of 6), a "1.5" generation (those who arrived in the US between the ages of 6-12), and a "1.25" generation (those who arrived in the US between the ages of 13-17). For these three groups, I consider every possible combination of coding such that earlier generations are not more acculturated than later generations (e.g. the 1.25 generation could not be "both" in cases where the 1.5 generation is "birthplace only"). Table 2 below shows the model fit of all ten possible combinations for both data sources.

```{r deviance-bendog}
dev_acs <- sapply(models_bendog_acs, function(x) {mean(x$deviance)}) %>%
  enframe(name="model", value="deviance_acs")

sapply(models_bendog_census, function(x) {mean(x$deviance)}) %>%
  enframe(name="model", value="deviance_census") %>%
  left_join(dev_acs) %>%
  mutate(gen1.75=c("Birthplace","USA","Both","USA","USA","USA","USA", "USA", 
                   "Both","Both"),
         gen1.5 =c("Birthplace","USA","Both","Both","Birthplace","USA","Both",
                   "USA", "Birthplace","Both"),
         gen1.25=c("Birthplace","USA","Both","Birthplace","Birthplace",
                   "Birthplace","Both","Both","Birthplace","Birthplace"),
         deviance_census=format(deviance_census, big.mark=","),
         deviance_acs=format(deviance_acs, big.mark=","),
         deviance_census=cell_spec(deviance_census,
                                   bold = ifelse(deviance_census=="1,122,736",
                                                       TRUE, FALSE)),
         deviance_acs=cell_spec(deviance_acs,
                                   bold = ifelse(deviance_acs=="1,849,202",
                                                       TRUE, FALSE)),
         ) %>%
  select(gen1.25, gen1.5, gen1.75, deviance_census, deviance_acs) %>%
  relocate(gen1.75, gen1.5, gen1.25, deviance_census, deviance_acs) %>%
  arrange(gen1.25, gen1.5, gen1.75) %>%
  kbl(caption = "Model fit to Census 1980 and ACS data using different specifications of birthplace endogamy for 1.25, 1.5, and 1.75 generations. Minimum deviance is shown in bold.", booktabs=T, longtable=F, 
      col.names=c("1.75","1.5","1.25","Census 1980","ACS 2014-18"),
      align = c("l","l","l","r","r"),
      linesep = NULL, escape = FALSE) %>%
  kable_styling(full_width = TRUE) %>%
  add_header_above(c("Generation" = 3, "Model Deviance" = 2)) %>%
  #column_spec(4, background = "red") %>%
  #column_spec(1, width = "5cm") %>%
# Goddmmit: apparently I need to use longtable to get wrapped footnotes, but then it won't float
  footnote(general=c("Age of US arrival by generation is 0-5 (1.75), 6-12 (1.5), and 13-17 (1.25). All models include",
                     "controls for educational differences, age differences, ethnoracial endogamy, and language",
                     "endogamy."), 
        threeparttable=F, footnote_as_chunk = F, 
          general_title="Notes: ", title_format="italic")
```

For both time period, the most preferred models use birthplace only coding for those respondents who entered the US after age 12. The USA only option representing full acculturation was not preferred for any group in either time period. In the Census 1980 data, the most preferred model treated all three groups the same as first generation individuals and only considered them endogamous with partners from the same actual birthplace. The more recent ACS data preferred a model in which both the 1.75 and 1.5 generations were considered birthplace endogamous both with partners from the actual birthplace or the USA. The shift over time suggests greater acculturation of the 1.5 and 1.75 generation today than in 1980. For all models in the analysis that follows, I code birthplace endogamy according to the best-fitting model for each data source from Table 2.

All models used here include controls for age and educational differences between potential partners. Age differences are modeled by taking the numerical age difference between partners and its square. Educational differences are modeled using educational crossing parameters [@schwartz_trends_2005]. All respondents are categorized as having less than a high school diploma, a high school diploma, some college education, or at least a four-year college degree. Three dummy variables indicate whether a particular union involves crossing the adjacent pairwise boundaries between these four categories. Results are cumulative such that a marriage involving one partner with less than a high school education and one partner with a four-year college degree would require passing all three boundaries.

# Results

## Birthplace and Language Diversity

Before moving to the model results, I first show the complex ways in which birthplace and language endogamy affect the likelihood of ethnic and pan-ethnic endogamy. I use the Blau index [@blau_inequality_1977] to measure language/birthplace diversity within each Asian and Latino ethnic group and to measure diversity within each pan-ethnic group. The Blau index of diversity ($D$) is calculated as: $$D=1-\sum_{i=1}^R p_i^2$$ where $p_i$ is the proportion of the total population belonging to group $i$ and $R$ is the total number of groups. The Blau index has an intuitively appealing interpretation as the probability that two randomly drawn members from the population belong to the same group.

```{r diversity-pan-bar, fig.cap="Birthplace and language diversity within Asian and Latino ethnic groups compared to birthplace and language diversity within pan-ethnic groups. Diversity is measured by the Blau index which gives the probability that two randomly selected members of the group do not share the same birthplace/language."}
ggplot(diversity_pan, aes(y=group, x=diversity, color=type))+
  #geom_col(position="dodge")+
  geom_vline(xintercept = 0, color="grey20", linetype=1)+
  geom_pointrangeh(aes(xmin=0, xmax=diversity), size=0.4,
                   position=position_dodgev(height=0.6))+
  facet_grid(variable~data)+
  scale_x_continuous(labels = scales::percent, breaks = c(0, 0.25, 0.5, 0.75))+
  scale_color_manual(values = my_palette[1:2])+
  labs(y=NULL, x="Probability two members do not share the same birthplace/language",
       fill="type of diversity")+
  theme_mine()+
  theme(panel.grid.major.y = element_blank())
```

Figure 2 shows the Blau index for language and birthplace across E/SE Asian, South Asian, Latino ethnic and pan-ethnic groups, using data on alternate partners in the two time periods. To simplify presentation, I take the mean diversity within each specific ethnic group and compare this directly to the diversity across the pan-ethnic group as a whole.

It is notable that there is significant language and birthplace diversity within specific ethnic groups (e.g. Chinese, Mexican). For example, the average diversity in birthplace within specific Latino ethnic groups in the ACS data is nearly 60%, indicating that two randomly determined members of the same ethnic group would not share the same birthplace about 60% of the time. This diversity is mostly driven by within ethnic group differences between being born in the place of origin or being born in the US and speaking the common language of the origin group or speaking English. Pan-ethnic diversity on the other hand is driven by additional differences in non-US birthplace and non-English language across specific ethnic groups. The larger the difference between within ethnic group diversity and pan-ethnic diversity, the larger the barrier there is to pan-ethnic intermarriage.

Figure 2 shows that pan-ethnic diversity in language and birthplace is greater than within ethnic group diversity for all three pan-ethnic groups in both time periods. However, these differences are trivially small Latinos, while they are quite large for Asians. Because Latinos share a common non-English language, this result was expected for language, but we also see a similar pattern here for birthplace. What differences exist among Latino ethnic groups have also diminished over time, whereas they have grown for E/SE Asians.

The consequence of these patterns is that language and birthplace endogamy is an important barrier to pan-ethnic intermarriage among Asians, but plays little role in the pan-ethnic intermarriage among Latinos. With that in mind, I now turn to models that estimate the likelihood of pan-ethnic intermarriage with and without accounting for birthplace and language endogamy.

## Modeling Pan-Ethnicity

Figure 3 shows the estimated odds of ethnic exogamy relative to ethnic endogamy for all three pan-ethnic groups in both time periods. As these odds ratios approach one, the odds of ethnic exogamy (e.g. Chinese person marrying a Japanese person) increase relative to ethnic endogamy (e.g. Japanese person marrying a Japanese person). Due to data limitations, results are only available for South Asians in the more recent ACS data.

```{r exog-lolly, fig.cap="The odds of ethnic exogamy for Latinos, East/Southeast Asians, and South Asians over time and model specification. Results are based on counterfactual marriage models. The baseline model controls for educational assortative mating and age differences."}
coefs_pent %>% 
  filter(variable=="E&SE Asian ethnic exogamy" | 
           variable=="Latino ethnic exogamy" |
           variable=="South Asian ethnic exogamy") %>%
  ggplot(aes(y=model, x=exp(coef), color=data))+
  geom_vline(xintercept = 0, color="grey20", linetype=1)+
  geom_vline(xintercept = 1, color="grey20", linetype=3)+
  geom_pointrangeh(aes(xmin=0, xmax=exp(coef)), size=0.4,
                   position=position_dodgev(height=0.6))+
  scale_color_manual(values=my_palette)+
  scale_x_continuous(limits = c(0,1))+
  labs(y=NULL, 
       x="Odds ratio of ethnic intermarriage relative to ethnic endogamy",
       color="Data source")+
  facet_wrap(~variable)+
  theme_mine()+
  theme(panel.spacing = unit(0.6, "lines"))
```

The baseline model includes only controls for age and educational differences between potential partners. I then include birthplace and language endogamy separately and together to determine the effects of these variables individually and collectively on the odds of ethnic exogamy.

In the baseline model, the odds of ethnic exogamy are similar for all three groups. In all cases, ethnic exogamy is about one-quarter as likely as ethnic endogamy. The results also show a very slight decline from the 1980 time period to the 2014-2018 when restricting to ethnic groups that were available in the 1980 data.

For Latinos, adjusting for birthplace and language endogamy has little effect on the odds ratio. Controlling for each variable individually increases the overall odds of ethnic exogamy only very slightly, with birthplac endogamy having a slightly larger effect. Overall, however, the odds are hardly changed at all. Furthermore, there is no evidence of change across the two time periods when limiting the analysis to comparable ethnic groups. In the later time period, ethnic exogamy is higher when including a larger range of ethnic groups, but still remains relatively low.

For both Asian pan-ethnic groups, controlling for birthplace and language endogamy substantially increases the odds of ethnic exogamy, to the point that ethnic exogamy in the later time period is not unusual. For example, using the more expansive ethnic group coding, the odds of ethnic exogamy relative to ethnic endogamy are roughly 0.75 to 1 for both E/SE Asians and South Asians once I control for both language and birthplace endogamy. Controlling for language endogamy produces larger changes than controlling for birthplace endogamy, but both produce substantial changes, individually and collectively.

Controlling for birthplace and language endogamy also reveals significant increases in the odds of ethnic exogamy over time for E/SE Asians. Using comparable ethnic groups, the odds of ethnic exogamy for E/SE Asians were about half those of ethnic endogamy in 1980, but have grown to 0.75 to 1 by the more recent time period.

Figure 3 shows the important role that birthplace and language endogamy play in pan-ethnic intermarriage. In actuality, we observe similar odds of ethnic exogamy among Asians and Latinos. However, the low odds of ethnic exogamy for Asians are largely a function of the high level of diversity in birthplace and language among Asians combined with a preference for marrying someone of the same birthplace and language. These barriers do not exist for Latinos. Controlling for these issues reveals very different affinities for pan-ethnicity among Latinos and Asians in the absence of birthplace and language barriers to pan-ethnic intermarriage.

```{r changes-intermar, fig.cap="Odds of ethnoracial exogamy relative to endogamy across two time periods. Results are based on models that control for age differences, educational differences, and birthplace and language endogamy. Values are sorted by ethnoracial exogamy in 1980. Arrows show the change across the two time periods based on comparable sets of ethnic groups. Results for American Indian/Alaska Native intermarriage are excluded due to sampling variability."}
coefs_pent %>%
  filter(model=="+ birthplace and\nlanguage endogamy" & 
           !grepl("South Asian ethnic exogamy", variable) & 
           !grepl("AIAN", variable)) %>%
  left_join(arrows_pent) %>%
  mutate(variable=str_replace(variable, "E&SE Asian ethnic exogamy", "**E&SE Asian ethnic exogamy**"),
         variable=str_replace(variable, "Latino ethnic exogamy", "**Latino ethnic exogamy**"),
         coef=ifelse(data=="ACS 2014-18, all ethnic groups" & variable=="White/Black", NA, coef)) %>%
  order_variables("Census 1980", "+ birthplace and\nlanguage endogamy") %>%
  ggplot(aes(x=variable, y=exp(coef)))+
  geom_hline(yintercept = 1, linetype=3, size=1)+
  geom_point(aes(color=data, shape=data), alpha=1, size=3)+
  geom_segment(aes(x=variable, xend=variable,
                   y=exp(coef.census), yend=exp(coef.acs)),
               color="grey30",
               arrow = arrow(length = unit(0.03, "npc")))+
  coord_flip()+
  scale_y_continuous(limits=c(0,1.1), expand=c(0.01,0))+
  scale_color_manual(values = my_palette)+
  scale_shape_manual(values = c(16,16,17))+
  labs(x=NULL, 
       y="Odds ratio of exogamy relative to ethnoracial endogamy",
       title=NULL,
       color="Data source", shape="Data source")+
  theme_mine()+
  theme(axis.text.y=element_markdown())
```

Figure 4 shows the odds of E/SE Asian and Latino ethnic exogamy in comparison to the odds of interracial marriage more broadly. Figure 4 also shows how these odds have changed over time. All results in Figure 4 are from the model that accounts for birthplace and language endogamy. E/SE Asian ethnic exogamy stands out as far more likely than any form of interracial marriage. Latino ethnic exogamy, on the other hand, is comparable in likelihood to several forms of interracial marriage. White/Latino intermarriage was less like likely than Latino ethnic exogamy in 1980, but has increased in frequency substantially. The consequence of this shift is that White/Latino exogamy is roughly as likely as Latino ethnic exogamy in the later time period (NOT QUITE TRUE FOR EXPANDED ETHNIC GROUPS).

Figure 4 also shows the results for South Asian interracial marriage over time. The results show significant declines in South Asian interracial marriage with all other groups over the time period. These results, however, should probably be viewed with caution. In the earlier time period, the only South Asian group available was "Asian Indian" and research has shown that respondents sometimes confuse the categories of "Asian Indian" and "American Indian" in race responses (CITE CAROLYN). Given the novel character of the "Asian Indian" category in 1980 (CHECK), it is likely that such misreporting has gone down over time. This may contribute to the overall decline in interracial marriage over time, if American Indians are less endogamous, on average, than Asian Indians.[^2] Nonetheless, the results here do not suggest any particular affinity between South Asians and E/SE Asians in either time period, suggesting that Asian pan-ethnicity occurs within two separate "melting pots."

[^2]: Although not shown here, I also found that the odds of intermarriage between American Indian/Alaska Natives and South Asians was slightly greater than one (EXACT NUMBER) in 1980 but declined to XX in the later ACS data. This result suggests that such race reporting errors played a greater role in 1980 than the later data.

Figure 4 also documents that the lowest likelihood of intermarriage in both time periods is that between blacks and all other groups, despite some substantial increases in these odds. These results indicate the enduring strength of the black/non-black boundary in the US despite increasing diversity in the overall population.

## Specific Affinities

The preceding models used a single ethnic exogamy term for each pan-ethnic group such that the odds of ethnic exogamy were assumed to be the same regardless of which specific groups were paired. For example, the odds of ethnic exogamy into a Japanese-Korean union would be the same as those for a Japanese-Filipino union. I now turn to models of the ACS data that relax this assumption and allow the estimation of specific affinities and disaffinities within the matrix of ethnic groups that make up the three pan-ethnic groups.

Fitting such models is actually quite difficult. If I considered every possible combination of ethnicities within pan-Ethnic groups in the ACS data, I would need to replace my three ethnic exogamy terms with 243 group specific parameters. Given that some of these groups are quite small in the dataset, models could not be fit to this fully expanded set, due to issues with sparseness and model convergence. I experimented with the largest possible set of ethnic groups for which I could reliably fit data and found that to include the five E/SE Asian ethnic groups available in the Census 1980 data (Chinese, Filipino, Vietnamese, Korean, and Japanese) and the ten largest Latino groups, with the exception of Hondurans. Because of the small size of non-Asian Indian groups in the South Asian categories I could not fit ethnicity specific models for South Asians.

In addition to separate parameters for each specific pairing of ethnic exogamy, I also include separate parameters for intermarriage with Whites and Blacks. These parameters allow me to examination variation in the tendency to intermarry with Whites and Blacks rather than ethnic exogamy for each specific ethnic group. 

Figure 5 shows the group-specific odds of ethnic exogamy for the five E/SE Asian ethnic groups, based on these models. I also provide a dendrogram that indicates the relative distance between each of these ethnic groups based on a tree diagram. The dendrogram is calculated by treating the inverse of the odds ratio as a measure of distance.

```{r asian-heat-map, fig.cap="Odds of ethnic exogamy relative to ethnic endogamy between particular East and Southeast Asian ethnic groups in ACS 2014-2018 data. Dendrogram on the right is based on hierarchical clustering treating the inverse of these odds ratios as a distance measure and using unweighted average distances. All models control for age differences, education differences, and language and birthplace endogamy."}
asian_groups <- c("Chinese","Korean","Japanese","Vietnamese","Filipino")
or <- coefs_extended %>%
  filter(data=="ACS 2014-18, expanded ethnic groups" &
           model=="+ both" &
           grepl(paste(asian_groups,collapse="|"), variable) &
           !grepl("Black|White", variable)) %>% #ugly hack, FIXME
  calc_or_matrix()
hc <- hclust(as.dist(or), method = "average")
or <- or[hc$order, rev(hc$order)]

gg.heat <- ggplot(melt(1/or, na.rm=TRUE), aes(Var2, Var1, fill = value))+
  geom_tile(color = NA)+
  scale_fill_gradient2(low = "#7b3294", high = "#008837", mid = "#f7f7f7", 
                       midpoint = 1, limit = c(0, 1.25),
                       name="odds ratio")+
  geom_text(aes(label=paste(round(1/or,2))))+
  labs(x=NULL, y=NULL)+
  theme_minimal()+
  theme(legend.position = "left", 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1))

dendro <- as.dendrogram(hc)
gg.dendro <- ggdendrogram(data = dendro, rotate = TRUE, labels = FALSE)+theme_void()

grid.newpage()
print(gg.heat, vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(gg.dendro, vp = viewport(x = 0.9, y = 0.57, width = 0.2, height = 0.68))
```

Figure 5 shows no social distance between the three East Asian groups of Chinese, Korean, and Japanese. The point estimates actually suggest that, once language and birthplace endogamy are accounted for, the odds of ethnic exogamy are slightly higher than ethnic endogamy among these groups. However, none of these odds ratios are statistically distinguishable from zero using a p-value cutoff of 0.05. 

Social distance remains between these East Asian groups and the Vietnamese and Filipino ethnic groups, with the exception of the Vietnamese/Chinese case where there is no barrier to ethnic exogamy. Filipinos, in particular, have the lowest odds of ethnic exogamy with  all of the other Asian groups, suggesting a stronger boundary separting Filipinos from wider E/SE Asian panethnicity. This may reflect the unique Spanish cultural inheritance of the Philippines. To more formally test this hypothesis, I included a specific dummy variable for Latino/Filipino intermarriage and the result was a large and statistically significant difference from the overall odds of E&SE Asian/Latino intermarriage. The odds of Latino/Filipino intermarriage was 0.28 to 1, in comparison to an odds of about 0.13 to 1 for intermarriage between Latinos and other E&SE Asian groups.

```{r latino-heat-map, fig.cap="Odds of ethnic exogamy relative to ethnic endogamy between particular Latino ethnic groups in ACS 2014-2018 data. Dendrogram on the right is based on hierarchical clustering treating the inverse of these odds ratios as a distance measure and using unweighted average distances. All models control for age differences, education differences, and language and birthplace endogamy."}
latino_groups <- c("Mexican","Puerto Rican","Cuban","Guatemalan",
            "Salvadorian","Colombian","Ecuadorian","Peruvian",
            "Dominican")
or <- coefs_extended %>%
  filter(data=="ACS 2014-18, expanded ethnic groups" &
           model=="+ both" &
           grepl(paste(latino_groups,collapse="|"), variable) &
           !grepl("Black|White", variable)) %>% #ugly hack, FIXME
  calc_or_matrix()
hc <- hclust(as.dist(or), method = "average")
or <- or[hc$order, rev(hc$order)]

gg.heat <- ggplot(melt(1/or, na.rm=TRUE), aes(Var2, Var1, fill = value))+
  geom_tile(color = NA)+
  scale_fill_gradient2(low = "#7b3294", high = "#008837", mid = "#f7f7f7", 
                       midpoint = 1, limit = c(0, 1.25),
                       name="odds ratio")+
  geom_text(aes(label=paste(round(1/or,2))), size=2.5)+
  labs(x=NULL, y=NULL)+
  theme_minimal()+
  theme(legend.position = "left", 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1))

dendro <- as.dendrogram(hc)
gg.dendro <- ggdendrogram(data = dendro, rotate = TRUE, labels = FALSE)+theme_void()

grid.newpage()
print(gg.heat, vp = viewport(x = 0.4, y = 0.5, width = 0.8, height = 1.0))
print(gg.dendro, vp = viewport(x = 0.9, y = 0.57, width = 0.2, height = 0.78))
```

Figure 6 shows comparable results for the nine largest Latino ethnic groups. Overall, the odds ratios in Figure 6 are much lower than those in Figure 5, reflecting the overall lower ethnic exogamy among Latino ethnic groups. The only case with no evidence of barriers to ethnic intermarriage was for Salvadorian and Guatemalans, where the estimated odds ratio is close to one.

The most notable feature of this figure is that ethnic affinities tend to cluster by region, with higher odds of intermarriage within the three groups of South American, Caribbean, and Mexican/Central American ethnicities. The one exception to this pattern is for Cubans, who tend to be about equal distance from both the South American and other Caribbean groups. 

Figure 7 shows the odds of intermarriage with Whites and Blacks for each of the E&SE Asian and Latino ethnic groups. for comparison, all of the ethnic exogamy odds ratios for a given ethnic group are also shown with small grey dots. 

```{r racial-exogamy, fig.cap="Odds of interracial marriage with Whites and Blacks for major Latino and E&SE Asian ethnic groups. For reference, odds of ethnic exogamy to other major ethnic groups are shown by small grey dots. Models are sorted from highest White/Black odds ratio to lowest. All models control for age differences, education differences, and language and birthplace endogamy."}
panethnic <- coefs_extended %>%
  filter(grepl(paste(c(latino_groups,asian_groups),collapse="|"), variable) & 
           !grepl("Black|White", variable) & 
           model=="+ both" & 
           data=="ACS 2014-18, expanded ethnic groups") %>%
  select(variable, coef)

#convert this to simple dataframe of all the coefs for each ethnic group
panethnic <- bind_rows(lapply(c(latino_groups,asian_groups), function(x) {
  return(tibble(ethnicity=x, 
                coef= panethnic$coef[str_detect(panethnic$variable, x)]))
}))
panethnic$group <- factor(panethnic$ethnicity %in% asian_groups,
                          levels=c(FALSE,TRUE), labels=c("Latino","E&SE ASian"))

intermar <- coefs_extended %>%
  filter(grepl(paste(c(latino_groups,asian_groups), collapse="|"), variable) & 
           grepl("Black|White", variable) & 
           model=="+ both" & 
           data=="ACS 2014-18, expanded ethnic groups") %>%
  mutate(ethnicity=str_remove(variable, "Black\\/|\\/White"),
         race=factor(str_detect(variable, "Black\\/"),
                     level=c(FALSE,TRUE), labels=c("White","Black")),
         group=factor(ethnicity %in% asian_groups,
                      levels=c(FALSE,TRUE), labels=c("Latino","E&SE ASian"))) %>%
  select(ethnicity, group, race, coef)

ggplot(intermar, 
       aes(y=reorder(ethnicity, coef, max), x=exp(coef)))+
  geom_vline(xintercept = 0, color="grey70", linetype=1)+
  geom_vline(xintercept = 1, color="grey70", linetype=3)+
  geom_point(aes(color=race), size=3)+
  geom_point(data=panethnic, alpha=0.5)+
  scale_color_manual(values=my_palette)+
  scale_x_continuous(limits=c(0,1.2))+
  facet_wrap(~group, scales = "free_y")+
  labs(y=NULL, 
       x="Odds ratio of interracial marriage relative to ethnic endogamy",
       color="Interracial partner")+
  theme_mine()+
  theme(legend.position = "bottom")
```

For all E&SE and Latino ethnic groups, intermarriage with Whites is more likely than intermarriage with Blacks. The variation in intermarriage trends across ethnic groups is more pronounced for Latino ethnic groups than for E&SE Asian groups. Among Latinos, Peruvians and Colombians are the most likely to intermarry with Whites with a odds ratios slightly higher than 0.6 to 1. At the other end of the spectrum the odds of Dominican/White intermarriage are slightly less than 0.25 to 1. For the five E&SE Asian groups, the odds ratios range from about 0.25 to 0.42.  

The odds of intermarriage with Blacks are uniformly low for all five E&SE Asian groups. The odds of intermarriage are also low for most Latino groups, with the exception of Dominicans and Puerto Ricans. For Dominicans, the odds of intermarriage with Blacks are almost identical to the odds of intermarriage with Whites. For Puerto Ricans, the odds of intermarriage with Blacks is the highest of any Latino ethnic group and is only slightly less likely than intermarriage with Whites. These differences almost certainly reflect the much higher prevalence of Afrodescent in Puerto Rican and Dominican populations and the racialization of members of these populations as black within the US (CITATIONS). However, this pattern is not observed for other Latino populations with a higher prevalence of Afrodescent, such as Cubans and Colombians. 

The relative placement of ethnic exogamy odds ratios for each ethnic group in Figure 7 is also telling. For all five Asian ethnic groups, every ethnic exogamy parameter is greater than the odds of intermarriage with Whites. For, the Latino groups, these ethnic exogamy parameters are frequently in between the odds of white and black intermarriage. For example, both Peruvians and Cubans are more likely to outmarry to a White person than to intermarry with any of the other Latino groups here. Mexicans are only slightly more likely to outmarry to most other Latino ethnic groups as they are to Whites, and are substantially less likely in two cases (Dominicans and Cubans). 

Overall, Figure 7 implies very different patterns of intermarriage for Latino and E&SE Asian ethnic groups. For the E&SE Asian ethnic groups, we observe only small variation across groups and all groups follow the same general pattern. Intermarriage with all other ethnic groups is more likely than intermarriage with Whites or Blacks, and the barriers to intermarriage with Blacks are far more substantial than the barriers to intermarriage with Whites. The consistency of this pattern speaks to a broad panethnic pattern of intermarriage for E&SE Asians. 

On the other hand, the results for Latino groups are characterized by ethnic-specific heterogeneity. In no case do we see a clear separation in which ethnic exogamy is preferred to outmarriage to Whites. We also see large differences in the tendency to outmarry with Whites and Blacks, both in the overall odds and in the difference in affinity. These results are not consistent with a broad tendency toward panethnicity among Latinos. Patterns of intermarriage instead seem to be very ethnically specific. 

# Conclusions

1.  Do we see evidence of pan-ethnicity in the marriage choices of Asians and Latinos and has this tendency changed over time?

2.  Answering this question is seriously complicated by the diversity of birthplace and language among and within Asian and Latino ethnic groups due to the continuing high levels of immigration from Asia and Latin America. The tendency for people to marry someone of the same birthplace and who speaks the same language is a significant barrier to pan-ethnic intermarriage for Asians, but not for Latinos.

3.  Once we account for birthplace and language endogamy, we see significant affinity between E/SE Asian and South Asian ethnic groups respectively. In the more recent ACS data, the odds of intermarrying with a member of a different ethnic group within the same pan-ethnic umbrella group were about 0.7 to 1 relative to marrying someone of the same exact ethnic group. In the case of E/SE Asians, where we can track this change over time, these odds have increased from about 0.5 to 1 in Census 1980. The results show little evidence of affinity between E/SE Asians and South Asians themselves, indicating that the growing pan-ethnicity among "Asians" is occurring in two distinct "melting pots."

4.  For Latinos, there is little evidence of change over time and the tendency toward ethnic exogamy is much lower than for Asians. Adjusting for birthplace and language endogamy, the odds of ethnic exogamy relative to ethnic endogamy for Latinos is roughly 0.25 to 1 in both time periods.

5.  It is important to recognize that these odds ratios reflect underlying affinities but not actual frequencies. In practice, the tendency toward ethnic exogamy is affected by birthplace and language diversity. This affect severely reduces the actual odds of ethnic exogamy for Asians much more than Latinos and produces odds ratios that are similar across all three groups, at about 0.25 to 1. Thus the higher underlying affinities among Asians would only be realized in a situation in which most Asians were US born and spoke English as a primary language, reflecting a situation of decreased immigration and greater acculturation.

6.  Given that we see no sign of declines in immigration, we are less likely to see clear pan-ethnic identities emerge from pan-ethnic intermarriage but rather what Massey has described as fractured identities among both Asians and Latinos.

7.  Why are the affinities among Latino ethnic groups lower than among Asian ethnic groups? This is somewhat surprising given that other research has noted the somewhat stronger basis for Latino pan-ethnicity. Language in particular has played a role in pan-ethnicity among Latinos, not simply in an ability to communicate in interpersonal situations, but in terms of creating pan-ethnic shared institutions. In some sense, language plays a role here in that it does not serve as a barrier to Latino groups in the same way that it does for Asians. However, the data clearly showed a weak affinity for ethnic exogamy among Latinos in comparison to Asians. This may reflect a deeper racialization of Asian identity. The idea of "Latino" as a racial category is still a contested idea and is contradicted by the phenotypic diversity of the Latino population, including but not limited to Afrodescent among some Latinos. On the other hand, "Asian" has a long history as a racial category in the US and more broadly, and was one of the original categories encoded by early racial taxonomic systems. In short, the experience of racialization may bind acculturated Asian Americans more closely together than ethnic identity. Latino identity, however, may remain more of a "federated" identity.

8.  Caveats about only one way to measure and identify pan-ethnicity.

# References
