---
title: "Changes in Pan-Ethnicity and Racial Boundaries in the United States based on Interracial Marriage Patterns, 1980-2018"
author:
- affiliation: University of Oregon, Sociology
  name: Aaron Gullickson
keywords: keywords
#FYI: the thanks and abstract are easier to view and edit in Atom which does proper linewrapping
thanks: Thanks to people and stuff
abstract: This is a test abstract
output:
  pdf_document: 
    citation_package: natbib
    fig_caption: yes
    template: ./resources/aog-latex-ms.tex
    keep_tex: no
  word_document:
    reference_docx: ./resources/aog_word_style.docx
fontfamily: mathpazo
fontsize: 11pt
anonymous: false #anonymize, double-space, and ragged right
endnotes: false #make footnotes endnotes
endfloat: false #move table and figures to the end
bibliography: ../project.bib
biblio-style: ./resources/ajs.bst
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, 
                      warning = FALSE, eval = TRUE,
                      fig.height=3.5)
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
load(here("analysis","output","models.RData"))
load(here("analysis","output","alternates.RData"))
load(here("analysis","output","models_bendog.RData"))
#you need to load kableExtra in the doc
options(kableExtra.latex.load_packages = FALSE)
options(knitr.table.format = "latex")
library(kableExtra)
library(ggtext)
library(stringr)

my_palette <- c("#7b3294","#008837","#a6dba0")

#instead of loading markets, which takes forever, hardcoding sample
#sizes but be sure to adjust if this changes
#load(here("analysis","output","markets_census.RData"))
#load(here("analysis","output","markets_acs_full.RData"))
#census_n <- sum(markets_census[[1]]$choice)
#acs_n <- sum(markets_acs_full[[1]]$choice)
census_n <- 285523
acs_n <- 503348

```

```{r theme, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, include=FALSE}
theme_mine <- function() {
  theme_bw()+
    theme(axis.title = element_text(size=10), 
          axis.text = element_text(size=8),
          legend.text=element_text(size=7),
          legend.title=element_text(size=8),
          legend.key.size = unit(0.8,"line"),
          strip.text = element_text(size=9),
          legend.position = "bottom")
}
```

```{r create-coef-table}
#extract coefficient values from model summary as tibble
get_coefs <- function(x) {
  temp <- coef(x)[,c(1,2)]
  return(tibble(variable=rownames(temp), coef=temp[,1], se=temp[,2]))
}
#bind together the model summaries
bind_models <- function(models_summary) {
  temp <- lapply(models_summary, get_coefs) %>%
    bind_rows(.id="model") %>%
    filter(grepl("race_exog", variable))
  return(temp)
}

## Pentagon Models ##
coefs_pent <- list(census=models_census_pentagon,
                    acs_1980basis=models_acs1980_pentagon,
                    acs_full=models_acsfull_pentagon) %>%
  lapply(bind_models) %>%
  bind_rows(.id="data") %>%
  mutate(variable=convert_intermar_names(variable, "race_exog_pent"),
         model=factor(model,
                      levels=c("base","bendog","lendog","both"),
                      labels=c("Baseline model", "+ birthplace\nendogamy",
                               "+ language\nendogamy", "+ birthplace and\nlanguage endogamy")),
         data=factor(data,
                     levels=c("census","acs_1980basis","acs_full"),
                     labels=c("Census 1980", 
                              "ACS 2014-18, 1980 ethnic groups",
                              "ACS 2014-18, all ethnic groups"))) %>%
  order_variables("Census 1980", "+ both")

#create a dataset that can be used for the arrows in the graph
#need a reworked dataset for arrows
arrows_pent <- coefs_pent %>%
  filter(model=="+ birthplace and\nlanguage endogamy" & 
           variable!="South Asian ethnic exogamy" &
           data!="ACS 2014-18, all ethnic groups") %>%
  select(variable, coef, data) %>% 
  pivot_wider(names_from=data, values_from=coef) %>%
  rename(coef.census=`Census 1980`, 
         coef.acs=`ACS 2014-18, 1980 ethnic groups`)# %>%
  #filter(abs(exp(coef.census)-exp(coef.acs))>0.05)

## Extended Models ##
coefs_extended <- list(census=models_census_extended,
                       acs_1980basis=models_acs1980_extended,
                       acs_full=models_acsres_extended) %>%
  lapply(bind_models) %>%
  bind_rows(.id="data") %>%
  mutate(variable=convert_intermar_names(variable, "race_exog_extended"),
         model=factor(model,
                      levels=c("base","bendog","lendog","both"),
                      labels=c("Baseline", "+ birthplace endogamy",
                               "+ language endogamy", "+ both")),
         data=factor(data,
                     levels=c("census","acs_1980basis","acs_full"),
                     labels=c("Census 1980", 
                              "ACS 2014-18, 1980 ethnic groups",
                              "ACS 2014-18, expanded ethnic groups"))) %>%
  order_variables("Census 1980", "+ both")

#arrows
arrows_extended <- coefs_extended %>%
  filter(model=="+ both" & 
           variable!="South Asian ethnic exogamy" &
           data!="ACS 2014-18, expanded ethnic groups") %>%
  select(variable, coef, data) %>% 
  pivot_wider(names_from=data, values_from=coef) %>%
  rename(coef.census=`Census 1980`, 
         coef.acs=`ACS 2014-18, 1980 ethnic groups`) %>%
  filter(abs(exp(coef.census)-exp(coef.acs))>0.05)
```

```{r diversity}
temp <- alternates_census %>%
  filter(race!="AIAN" & race!="White" & race!="Black") %>%
  mutate(group=case_when(
    race %in% c("Chinese","Korean","Japanese","Vietnamese",
                     "Filipino","Cambodian","Hmong","Laotian",
                     "Thai","Burmese","Indonesian","Malaysian") ~ "E&SE Asian",
    race %in% c("Asian Indian","Bangladeshi",
                     "Pakistani","Sri Lankan") ~ "South Asian",
    race %in% c("Mexican","Puerto Rican","Cuban","Costa Rican",
                "Guatemalan","Honduran","Nicaraguan","Panamanian",
                "Salvadorian","Argentinian","Bolivian","Chilean",
                "Colombian","Ecuadorian","Paraguayan","Peruvian",
                "Uruguayan","Venezuelan","Dominican") ~ "Latino"),
    group=factor(group, levels=c("E&SE Asian","South Asian","Latino")))

diversity_ethnic <- temp %>%
  count(race, group, bpld) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980")

diversity_ethnic <- temp %>%
  count(race, group, lang) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980") %>%
  full_join(diversity_ethnic)

diversity_pan <- temp %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980")

diversity_pan <- temp %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980") %>%
  full_join(diversity_pan)

temp <- alternates_acs %>%
  filter(race!="AIAN" & race!="White" & race!="Black") %>%
  mutate(group=case_when(
    race %in% c("Chinese","Korean","Japanese","Vietnamese",
                     "Filipino","Cambodian","Hmong","Laotian",
                     "Thai","Burmese","Indonesian","Malaysian") ~ "E&SE Asian",
    race %in% c("Asian Indian","Bangladeshi",
                     "Pakistani","Sri Lankan") ~ "South Asian",
    race %in% c("Mexican","Puerto Rican","Cuban","Costa Rican",
                "Guatemalan","Honduran","Nicaraguan","Panamanian",
                "Salvadorian","Argentinian","Bolivian","Chilean",
                "Colombian","Ecuadorian","Paraguayan","Peruvian",
                "Uruguayan","Venezuelan","Dominican") ~ "Latino"),
    group=factor(group, levels=c("E&SE Asian","South Asian","Latino")))

temp2 <- temp %>%
  count(race, group, bpld) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18")

diversity_ethnic <- temp %>%
  count(race, group, lang) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18") %>%
  full_join(temp2) %>%
  bind_rows(diversity_ethnic) %>%
  select(race, group, data, diversity_bpl, diversity_lang)

temp2 <- temp %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\nall ethnic groups")

diversity_pan <- temp %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\nall ethnic groups") %>%
  full_join(temp2) %>%
  bind_rows(diversity_pan)

races1980 <- as.character(subset(diversity_ethnic, data=="Census 1980")$race)

temp2 <- temp %>%
  filter(race %in% races1980) %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups")

diversity_pan <- temp %>%
  filter(race %in% races1980) %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups") %>%
  full_join(temp2) %>%
  bind_rows(diversity_pan)  %>%
  mutate(type="pan-ethnic group diversity")

temp2 <- diversity_ethnic %>%
  group_by(group, data) %>%
  summarize(diversity_bpl=mean(diversity_bpl),
            diversity_lang=mean(diversity_lang)) %>%
  ungroup() %>%
  mutate(data=ifelse(data=="ACS 2014-18", "ACS 2014-18\nall ethnic groups",
                     data))

diversity_pan <- diversity_ethnic %>%
  filter(race %in% races1980 & data=="ACS 2014-18") %>%
  group_by(group, data) %>%
  summarize(diversity_bpl=mean(diversity_bpl),
            diversity_lang=mean(diversity_lang)) %>%
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups") %>%
  bind_rows(temp2)  %>%
  mutate(type="mean of within ethnic group diversity") %>%
  bind_rows(diversity_pan) %>%
  pivot_longer(cols=starts_with("diversity_"), names_to="variable",
               values_to="diversity") %>%
  mutate(variable=ifelse(variable=="diversity_bpl",
                     "Birthplace",
                     "Language"),
         diversity=ifelse(group=="South Asian" & 
                            data!="ACS 2014-18\nall ethnic groups",
                          NA, diversity),
         data=factor(data,
                     levels=c("Census 1980", 
                              "ACS 2014-18\n1980 ethnic groups",
                              "ACS 2014-18\nall ethnic groups")),
         group=factor(group, 
                      levels=c("E&SE Asian", "Latino","South Asian")))
```

# Introduction

# Background

# Data and Methods

The data for this analysis are derived from the microdata sample of the 1980 US Census and the American Community Survey (ACS) for 2014-2018. The ACS is an annual 1-in-100 survey of the United States population, conducted by the United States Census Bureau. The 1980 US Census is the last Census to include information about marriage timing. Information about marriage timing was added to the ACS in 2008. To increase sample size for smaller ethnoracial categories, I pool five years of ACS data from 2014-2018. Both data sources were extracted from the Integrated Public Use Microdata Series (IPUMS) system [@ruggles_ipums_2020].

I use a counterfactual marriage model [@qian_marriage_2018; @gullickson_counterfactual_2021] for the analysis that follows. In both data sources, I restrict the analysis to all opposite-sex marriages formed in the previous five years that were a first marriage for both partners. The restriction to opposite-sex first marriages is necessary for comparison because the Census 1980 only recorded marriage timing relative to first marriage and does not include same-sex unions. For each marriage, I then construct a choice set of one real union and twenty fictional unions. Fictional unions are created by sampling alternate partners for one randomly determined spouse from a pool of potential partners. I then use a conditional logit model to predict how partner characteristics relate to the likelihood of observing the true union, as follows:
$$P_{ij}=\frac{e^{\mathbf{x}_{ij}\beta}}{\sum_{k=1}^J e^{\mathbf{x}_{ik}\beta}}$$
where $P_{ij}$ is the probability that union $j$ within choice set $i$ is the actual union. $J$ is the total number of unions in the choice set (in this case, twenty-one). The vector $\mathbf{x}_{ij}$ defines the characteristics of the union and the $\beta$ vector provides estimated log-odds ratios indicating how the odds of an actual union change with $\mathbf{x}_{ij}$. The model is estimated as a fixed-effects logistic regression model with fixed effects for each choice set. 

The set of potential alternate partners from which fictional spouses are sampled is made up of all unmarried adults as well as those who were married in the previous five years. Because of uncertainty regarding timing of migration and marriage, individuals who migrated to the US or across state lines in the last five years are excluded from analysis as both actual and alternate marriage partners. To better capture the supply of potential partners in local marriage markets, alternate partners are sampled from among individuals residing in the same marriage market. I define marriage markets by the metropolitan area of residence when this can be identified for a respondent, and otherwise by the state of residence. Metropolitan area is not directly identifiable in the data used here due to confidentiality concerns, but can be deduced for many metropolitan areas based on the finest geographic identifier of the respondent (county group in Census 1980 and Public Use Microdata Area in the ACS). IPUMS provides metro area identification for respondents in cases where the county group or PUMA was mostly contained within the metropolitan area. Figure 1 shows the identifiable metropolitan areas used in the analysis for both time periods. Although not all metropolitan areas are identified in the data, most major metropolitan areas are included in both time periods. Coverage differs somewhat by period, with more extensive coverage in the later time period. 

```{r, fig.cap="Metropolitan statistical areas identifiable in the data by data source. Fill colors are drawn with partial transparency so that darker areas indicate overlapping areas in the two data sources."}
knitr::include_graphics(here("analysis","output","msa_map.pdf"), 
                        auto_pdf = TRUE)
```



Because this approach relies upon a random sample of alternate partners, results will vary each time this sampling procedure is performed. To account for the added uncertainty due to sampling, I preform the sampling procedure separately five times to produce five separate analytical datasets. I then run parallel models on all five datasets and pool $\beta$ estimates and standard errors using the same methods as those employed for multiple imputation [@rubin_multiple_1987]. 

This counterfactual marriage model approach has several advantages over the more well known log-linear approach to analyzing assortative mating. Most importantly, because the researcher can specify the parameters for sampling alternate partners, it is relatively easy to adjust for sub-national differences in group composition. Smaller groups in the population are more likely to be exogamous simply because there are fewer within-group potential partners available. Assortative mating researchers typically desire to factor out such "supply side" issues when estimating the likelihood of group exogamy. @harris_how_2005 have demonstrated that the uneven spatial distribution of racial groups across the US plays a significant role in restricting racial exogamy and log-linear models estimated at the national level will not account for this unevenness. This issue is likely to play an important role in this analysis, because Asian and Latino ethnic groups are not evenly distributed across the US. I account for this uneven spatial distribution by sampling alternate partners from within the same metropolitan area (or state, for respondents outside of an identifiable metropolitan area) such that potential partners more closely match the composition of the local marriage market for the focal partner.

The counterfactual model also has two other notable advantages over the log-linear model. First, it includes the unmarried population within the sample of alternate partners whereas the log-linear model only includes those who were married. Second, the linear model framework makes incorporation of a variety of quantitative and categorical control variables much easier.

## Variable Construction

I measure ethnoracial categories by a combination of the race and Hispanicity questions in both data sources. Table 1 shows the sample size for all ethnoracial groups included in the analysis. I classify non-Asian and non-Latino respondents as White, Black, or American Indian/Alaska Native (AIAN). Due to small sample size, I exclude Pacific Islanders, multriacial respondents, and non-Latino others from the analysis. The race question includes several Asian nationalities (e.g. Chinese, Korean, Filipino) as well as a write-in response for ethnic identities not captured by the existing categories. Similarly, the Hispanicity question includes major Latino ethnic groups (e.g. Mexican, Puerto Rican, Cuban) as well as a write-in option. Importantly, the Census 1980 data includes a far more limited set of identifiable ethnic groups for both Asians and Latinos than the ACS data. In Census 1980, I can only identify Mexicans, Puerto Ricans, and Cubans among Latinos and Chinese, Japanese, Korean, Filipino, Vietnamese, and Asian Indian respondents among Asians. To address this restriction, I estimate two different kinds of models for the ACS data. In the first model, I restrict the data only to those ethnic groups that were available in the Census 1980 data to allow more direct comparison over time. I also estimate models using the full ACS data, but these are not directly comparable to the Census 1980 results.

```{r sample-size-table}

temp <- c(table(alternates_acs$race)) %>% enframe()
tab_samp <- c(table(alternates_census$race)) %>% 
  enframe() %>%
  right_join(temp, by="name") %>%
  rename(ethnorace=name, census=value.x, acs=value.y) %>%
  mutate(race_big=case_when(
    ethnorace=="Chinese" ~ "East and Southeast Asian",
    ethnorace=="Japanese" ~ "East and Southeast Asian",
    ethnorace=="Korean" ~ "East and Southeast Asian",
    ethnorace=="Filipino" ~ "East and Southeast Asian",
    ethnorace=="Vietnamese" ~ "East and Southeast Asian",
    ethnorace=="Cambodian" ~ "East and Southeast Asian",
    ethnorace=="Hmong" ~ "East and Southeast Asian",
    ethnorace=="Laotian" ~ "East and Southeast Asian",
    ethnorace=="Thai" ~ "East and Southeast Asian",
    ethnorace=="Burmese" ~ "East and Southeast Asian",
    ethnorace=="Indonesian" ~ "East and Southeast Asian",
    ethnorace=="Malaysian" ~ "East and Southeast Asian",
    ethnorace=="Bhutanese" ~ "East and Southeast Asian",
    ethnorace=="Nepalese" ~ "East and Southeast Asian",
    ethnorace=="Asian Indian"  ~ "South Asian",
    ethnorace=="Pakistani" ~ "South Asian",
    ethnorace=="Bangladeshi" ~ "South Asian",
    ethnorace=="Sri Lankan" ~ "South Asian",
    ethnorace=="Mexican" ~ "Latino", 
    ethnorace=="Cuban" ~ "Latino", 
    ethnorace=="Puerto Rican" ~ "Latino", 
    ethnorace=="Guatemalan" ~ "Latino", 
    ethnorace=="Salvadorian" ~ "Latino", 
    ethnorace=="Dominican" ~ "Latino", 
    ethnorace=="Colombian" ~ "Latino", 
    ethnorace=="Costa Rican" ~ "Latino", 
    ethnorace=="Honduran" ~ "Latino", 
    ethnorace=="Nicaraguan" ~ "Latino", 
    ethnorace=="Panamanian" ~ "Latino", 
    ethnorace=="Argentinian" ~ "Latino", 
    ethnorace=="Bolivian" ~ "Latino", 
    ethnorace=="Chilean" ~ "Latino", 
    ethnorace=="Ecuadorian" ~ "Latino", 
    ethnorace=="Peruvian" ~ "Latino", 
    ethnorace=="Venezuelan" ~ "Latino", 
    ethnorace=="Paraguayan" ~ "Latino", 
    ethnorace=="Uruguayan"  ~ "Latino"
  ))

temp <- tab_samp %>% group_by(race_big) %>%
  summarize(census=sum(census, na.rm=TRUE), acs=sum(acs, na.rm=TRUE)) %>%
  na.omit() %>%
  rename(ethnorace=race_big)

tab_samp <- tab_samp %>% select(ethnorace, census, acs) %>%
  bind_rows(temp)

temp <- tibble(ethnorace="Marriages in the previous year",
               census=census_n, acs=acs_n)

tab_samp <- tab_samp %>% 
  bind_rows(temp)
                    
ord <- c("Marriages in the previous year",
         "White","Black","AIAN","Latino",
         "Mexican","Puerto Rican","Cuban","Salvadorian","Dominican","Guatemalan",
         "Colombian","Honduran","Peruvian","Ecuadorian","Nicaraguan","Argentinian",
         "Venezuelan","Panamanian","Chilean","Costa Rican","Bolivian",
         "Uruguayan","Paraguayan",
         "East and Southeast Asian",
         "Chinese","Filipino","Vietnamese","Korean","Japanese","Cambodian",
         "Hmong","Laotian","Thai","Burmese","Indonesian","Malaysian",
         "South Asian",
         "Asian Indian","Pakistani","Bangladeshi","Sri Lankan")

tab_samp <- as.data.frame(tab_samp)
rownames(tab_samp) <- tab_samp$ethnorace
tab_samp <- tab_samp[ord,]
rownames(tab_samp) <- NULL

options(knitr.kable.NA = '')

tab_samp %>%
  mutate(census=format(census, big.mark=","),
         census=ifelse(census=="       NA","",census),
         acs=format(acs, big.mark=",")) %>%
  kbl(caption = "Sample size of marriages and alternate partners by data source and ethnoracial category.", booktabs=T, longtable=F,
      linesep = NULL,
      col.names=c("Category","Census 1980", "ACS 2014-2018"),
      align = c("l","r","r")) %>%
  kable_styling(full_width = TRUE) %>%
  add_indent(c(6:24,26:37,39:42)) %>%
  pack_rows("Marriages", 1, 1) %>%
  pack_rows("Alternate Partners", 2, 42) %>%
  column_spec(1, width = "6cm")
```

I model interracial and panethnic marriage using a set of dummy variables where the reference category is a racially and ethnically endogamous union. Because the number of ethnoracial categories is large, modeling every possible combination of these categories for exogamous unions is not feasible, so I apply two limits on the parameterization. First, when analyzing intermarriage with Whites, Blacks, and American Indians/Alaska Natives, I group Asian and Latino ethnic groups into the larger pan-ethnic categories of Latino, East/Southeast Asian, and South Asian. For example, a union involving a black and Mexican partner and a union involving a black and Colombian partner would both be coded as a Black/Latino union. Second, for most of the models that follow, I model the likelihood of ethnic exogamy using a single dummy variable to indicate any form of ethnic exogamy within the three pan-ethnic categories of Latino, East/Southeast Asian, and South Asian. This allows me to estimate the average likelihood of ethnic exogamy within these groups, but does not capture particular affinities of disaffinities among ethnic groups within the pan-ethnic category. In some of the models that follow, I relax this assumption for the largest ethnic groups in order to estimate such affinities and disaffinities in the more recent ACS data.

Because I separate South Asian from East/Southeast Asian groups and only one South Asian ethnic group was identifiable in the Census 1980 data, I cannot estimate pan-ethnic parameters over time for South Asians. Thus, most of the analysis that follows will focus on pan-ethnicity among East/Southeast Asians and Latinos. However, I do estimate and report pan-ethnic parameters for South Asians for some models using ACS data.

I account for language and birthplace endogamy with simple dummy variables indicating whether the two potential partners share the same primary language or birthplace, respectively. Primary language is determined by what language the respondent reported speaking at home. Because this variable is measured after a marriage has occurred, it may somewhat overestimate language endogamy among respondents. Some partners who spoke different languages when they met may have shifted the language spoken at home after marriage. 

Birthplace endogamy is complicated by the "1.5" generation -- individuals who were born outside of the United States but migrated as children and whose formative experiences are at least partially defined by acculturation within the US. Within the coding scheme used here, I consider three possibilities for such individuals organized by the assumed degree of acculturation and assimilation. First such individuals could be considered birthplace endogamous only with individuals from the same actual birthplace. Second, such individuals could be considered birthplace endogamous with either an individual from their actual birthplace or the a person born in the USA. Third, such individuals could be considered only birthplace endogamous with other individuals born in the USA. 

To test the accuracy of these three scenarios, I fit models using each coding scheme to the Census 1980 and full ACS data. Following @rumbaut_ages_2004a, I also divide the 1.5 generation into a "1.75" generation (those who arrived in the US before the age of 6), a "1.5" generation (those who arrived in the US between the ages of 6-12), and a "1.25" generation (those who arrived in the US between the ages of 13-17). For these three groups, I consider every possible combination of coding such that earlier generations are not more acculturated than later generations (e.g. the 1.25 generation could not be "both" in cases where the 1.5 generation is "birthplace only"). Table 2 below shows the model fit of all ten possible combinations for both data sources.

```{r deviance-bendog}
dev_acs <- sapply(models_bendog_acs, function(x) {mean(x$deviance)}) %>%
  enframe(name="model", value="deviance_acs")

sapply(models_bendog_census, function(x) {mean(x$deviance)}) %>%
  enframe(name="model", value="deviance_census") %>%
  left_join(dev_acs) %>%
  mutate(gen1.75=c("Birthplace","USA","Both","USA","USA","USA","USA", "USA", 
                   "Both","Both"),
         gen1.5 =c("Birthplace","USA","Both","Both","Birthplace","USA","Both",
                   "USA", "Birthplace","Both"),
         gen1.25=c("Birthplace","USA","Both","Birthplace","Birthplace",
                   "Birthplace","Both","Both","Birthplace","Birthplace"),
         deviance_census=format(deviance_census, big.mark=","),
         deviance_acs=format(deviance_acs, big.mark=","),
         deviance_census=cell_spec(deviance_census,
                                   bold = ifelse(deviance_census=="1,122,736",
                                                       TRUE, FALSE)),
         deviance_acs=cell_spec(deviance_acs,
                                   bold = ifelse(deviance_acs=="1,849,202",
                                                       TRUE, FALSE)),
         ) %>%
  select(gen1.25, gen1.5, gen1.75, deviance_census, deviance_acs) %>%
  relocate(gen1.75, gen1.5, gen1.25, deviance_census, deviance_acs) %>%
  arrange(gen1.25, gen1.5, gen1.75) %>%
  kbl(caption = "Model fit to Census 1980 and ACS data using different specifications of birthplace endogamy for 1.25, 1.5, and 1.75 generations. Minimum deviance is shown in bold.", booktabs=T, longtable=F, 
      col.names=c("1.75","1.5","1.25","Census 1980","ACS 2014-18"),
      align = c("l","l","l","r","r"),
      linesep = NULL, escape = FALSE) %>%
  kable_styling(full_width = TRUE) %>%
  add_header_above(c("Generation" = 3, "Model Deviance" = 2)) %>%
  #column_spec(4, background = "red") %>%
  #column_spec(1, width = "5cm") %>%
# Goddmmit: apparently I need to use longtable to get wrapped footnotes, but then it won't float
  footnote(general=c("Age of US arrival by generation is 0-5 (1.75), 6-12 (1.5), and 13-17 (1.25). All models include",
                     "controls for educational differences, age differences, ethnoracial endogamy, and language",
                     "endogamy."), 
        threeparttable=F, footnote_as_chunk = F, 
          general_title="Notes: ", title_format="italic")
```

For both time period, the most preferred models use birthplace only coding for those respondents who entered the US after age 12. The USA only option representing full acculturation was not preferred for any group in either time period. In the Census 1980 data, the most preferred model treated all three groups the same as first generation individuals and only considered them endogamous with partners from the same actual birthplace. The more recent ACS data preferred a model in which both the 1.75 and 1.5 generations were considered birthplace endogamous both with partners from the actual birthplace or the USA. The shift over time suggests greater acculturation of the 1.5 and 1.75 generation today than in 1980. For all models in the analysis that follows, I code birthplace endogamy according to the best-fitting model for each data source from Table 2. 

All models used here include controls for age and educational differences between potential partners. Age differences are modeled by taking the numerical age difference between partners and its square. Educational differences are modeled using educational crossing parameters [@schwartz_trends_2005]. All respondents are categorized as having less than a high school diploma, a high school diploma, some college education, or at least a four-year college degree. Three dummy variables indicate whether a particular union involves crossing the adjacent pairwise boundaries between these four categories. Results are cumulative such that a marriage involving one partner with less than a high school education and one partner with a four-year college degree would require passing all three boundaries.

# Results

## Birthplace and Language Diversity


```{r diversity-pan-bar, fig.cap="Birthplace and language diversity within Asian and Latino ethnic groups compared to birthplace and language diversity within pan-ethnic groups. Diversity is measured by the Gini-Simpson index which gives the probability that two randomly selected members of the group do not share the same birthplace/language."}
ggplot(diversity_pan, aes(y=group, x=diversity, color=type))+
  #geom_col(position="dodge")+
  geom_pointrangeh(aes(xmin=0, xmax=diversity), size=0.4,
                   position=position_dodgev(height=0.6))+
  facet_grid(variable~data)+
  scale_x_continuous(labels = scales::percent, breaks = c(0, 0.25, 0.5, 0.75))+
  scale_color_manual(values = my_palette[1:2])+
  labs(y=NULL, x="Probability two members do not share the same birthplace/language",
       fill="type of diversity")+
  theme_mine()+
  theme(panel.grid.major.y = element_blank())
```


## Modeling Pan-Ethnicity

```{r exog-lolly, fig.cap="The odds of ethnic exogamy for Latinos, East/Southeast Asians, and South Asians over time and model specification. Results are based on counterfactual marriage models. The baseline model controls for educational assortative mating and age differences."}
coefs_pent %>% 
  filter(variable=="E&SE Asian ethnic exogamy" | 
           variable=="Latino ethnic exogamy" |
           variable=="South Asian ethnic exogamy") %>%
  ggplot(aes(y=model, x=exp(coef), color=data))+
  geom_vline(xintercept = 0, color="grey20", linetype=1)+
  geom_vline(xintercept = 1, color="grey20", linetype=3)+
  geom_pointrangeh(aes(xmin=0, xmax=exp(coef)), size=0.4,
                   position=position_dodgev(height=0.6))+
  scale_color_manual(values=my_palette)+
  scale_x_continuous(limits = c(0,1))+
  labs(y=NULL, 
       x="Odds ratio of ethnic intermarriage relative to ethnic endogamy",
       color="Data source")+
  facet_wrap(~variable)+
  theme_mine()+
  theme(panel.spacing = unit(0.6, "lines"))
```

```{r changes-intermar, fig.cap="Odds of ethnoracial exogamy relative to endogamy across two time periods. Results are based on models that control for age differences, educational differences, and birthplace and language endogamy. Values are sorted by ethnoracial exogamy in 1980. Arrows show the change across the two time periods based on comparable sets of ethnic groups. Results for American Indian/Alaska Native intermarriage are excluded due to sampling variability."}
coefs_pent %>%
  filter(model=="+ birthplace and\nlanguage endogamy" & 
           !grepl("South Asian ethnic exogamy", variable) & 
           !grepl("AIAN", variable)) %>%
  left_join(arrows_pent) %>%
  mutate(variable=str_replace(variable, "E&SE Asian ethnic exogamy", "**E&SE Asian ethnic exogamy**"),
         variable=str_replace(variable, "Latino ethnic exogamy", "**Latino ethnic exogamy**"),
         coef=ifelse(data=="ACS 2014-18, all ethnic groups" & variable=="White/Black", NA, coef)) %>%
  order_variables("Census 1980", "+ birthplace and\nlanguage endogamy") %>%
  ggplot(aes(x=variable, y=exp(coef)))+
  geom_hline(yintercept = 1, linetype=3, size=1)+
  geom_point(aes(color=data, shape=data), alpha=1, size=3)+
  geom_segment(aes(x=variable, xend=variable,
                   y=exp(coef.census), yend=exp(coef.acs)),
               color="grey30",
               arrow = arrow(length = unit(0.03, "npc")))+
  coord_flip()+
  scale_y_continuous(limits=c(0,1.1), expand=c(0,0))+
  scale_color_manual(values = my_palette)+
  scale_shape_manual(values = c(16,16,17))+
  labs(x=NULL, 
       y="Odds ratio of exogamy relative to ethnoracial endogamy",
       title=NULL,
       color="Data source", shape="Data source")+
  theme_mine()+
  theme(axis.text.y=element_markdown())
```

What other figures
* A two panel figure showing tiles/trees for specific ethnicity models
  * or alternatively two split panel figures, one showing change over time in these characteristics and the other showing tile, and then separate figs for E&SE Asians and Latinos
* Black/Latino exogamy? One panel showing change over time and the other showing odds ratios in ACS data

# Conclusions

