---
title: "Patterns of Panthnic Intermarriage in the United States, 1980-2018"
author:
- affiliation: University of Oregon, Sociology
  name: Aaron Gullickson
keywords: keywords
#FYI: the thanks and abstract are easier to view and edit in Atom which does proper linewrapping
thanks: Thanks to people and stuff
abstract: This is a test abstract
output:
  bookdown::pdf_document2: 
    citation_package: natbib
    fig_caption: yes
    template: ./resources/aog-latex-ms.tex
    keep_tex: no
    toc: false
  word_document:
    reference_docx: ./resources/aog_word_style.docx
fontfamily: mathpazo
fontsize: 11pt
anonymous: false #anonymize, double-space, and ragged right
endnotes: false #make footnotes endnotes
endfloat: false #move table and figures to the end
bibliography: ../project.bib
biblio-style: ./resources/ajs.bst
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, error = FALSE, 
                      warning = FALSE, eval = TRUE,
                      fig.height=3.5)
library(here)
source(here("analysis","check_packages.R"))
source(here("analysis","useful_functions.R"))
load(here("analysis","output","models.RData"))
load(here("analysis","output","alternates.RData"))
load(here("analysis","output","models_bendog.RData"))
#you need to load kableExtra in the doc
options(kableExtra.latex.load_packages = FALSE)
options(knitr.table.format = "latex")
library(kableExtra)
library(ggtext)
library(stringr)

my_palette <- c("#7b3294","#008837","#a6dba0")

#instead of loading markets, which takes forever, hardcoding sample
#sizes but be sure to adjust if this changes
#load(here("analysis","output","markets_census.RData"))
#load(here("analysis","output","markets_acs_full.RData"))
#census_n <- sum(markets_census[[1]]$choice)
#acs_n <- sum(markets_acs_full[[1]]$choice)
census_n <- 285523
acs_n <- 503348

```

```{r theme, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, include=FALSE}
theme_mine <- function() {
  theme_bw()+
    theme(axis.title = element_text(size=10), 
          axis.text = element_text(size=8),
          legend.text=element_text(size=7),
          legend.title=element_text(size=8),
          legend.key.size = unit(0.8,"line"),
          strip.text = element_text(size=9),
          legend.position = "bottom")
}
```

```{r create-coef-table}
#extract coefficient values from model summary as tibble
get_coefs <- function(x) {
  temp <- coef(x)[,c(1,2)]
  return(tibble(variable=rownames(temp), coef=temp[,1], se=temp[,2]))
}
#bind together the model summaries
bind_models <- function(models_summary) {
  temp <- lapply(models_summary, get_coefs) %>%
    bind_rows(.id="model") %>%
    filter(grepl("race_exog", variable))
  return(temp)
}

## Pentagon Models ##
coefs_pent <- list(census=models_census_pentagon,
                    acs_1980basis=models_acs1980_pentagon,
                    acs_full=models_acsfull_pentagon) %>%
  lapply(bind_models) %>%
  bind_rows(.id="data") %>%
  mutate(variable=convert_intermar_names(variable, "race_exog_pent"),
         model=factor(model,
                      levels=c("base","bendog","lendog","both"),
                      labels=c("Baseline model", "+ birthplace\nendogamy",
                               "+ language\nendogamy", "+ birthplace and\nlanguage endogamy")),
         data=factor(data,
                     levels=c("census","acs_1980basis","acs_full"),
                     labels=c("Census 1980", 
                              "ACS 2014-18, 1980 ethnic groups",
                              "ACS 2014-18, all ethnic groups"))) %>%
  order_variables("Census 1980", "+ both")

#create a dataset that can be used for the arrows in the graph
#need a reworked dataset for arrows
arrows_pent <- coefs_pent %>%
  filter(model=="+ birthplace and\nlanguage endogamy" & 
           variable!="South Asian ethnic exogamy" &
           data!="ACS 2014-18, all ethnic groups") %>%
  select(variable, coef, data) %>% 
  pivot_wider(names_from=data, values_from=coef) %>%
  rename(coef.census=`Census 1980`, 
         coef.acs=`ACS 2014-18, 1980 ethnic groups`)# %>%
  #filter(abs(exp(coef.census)-exp(coef.acs))>0.05)

## Extended Models ##
coefs_extended <- list(census=models_census_extended,
                       acs_1980basis=models_acs1980_extended,
                       acs_full=models_acsres_extended) %>%
  lapply(bind_models) %>%
  bind_rows(.id="data") %>%
  mutate(variable=convert_intermar_names(variable, "race_exog_extended"),
         model=factor(model,
                      levels=c("base","bendog","lendog","both"),
                      labels=c("Baseline", "+ birthplace endogamy",
                               "+ language endogamy", "+ both")),
         data=factor(data,
                     levels=c("census","acs_1980basis","acs_full"),
                     labels=c("Census 1980", 
                              "ACS 2014-18, 1980 ethnic groups",
                              "ACS 2014-18, expanded ethnic groups"))) %>%
  order_variables("Census 1980", "+ both")

#arrows
arrows_extended <- coefs_extended %>%
  filter(model=="+ both" & 
           variable!="South Asian ethnic exogamy" &
           data!="ACS 2014-18, expanded ethnic groups") %>%
  select(variable, coef, data) %>% 
  pivot_wider(names_from=data, values_from=coef) %>%
  rename(coef.census=`Census 1980`, 
         coef.acs=`ACS 2014-18, 1980 ethnic groups`) %>%
  filter(abs(exp(coef.census)-exp(coef.acs))>0.05)
```

```{r diversity}
temp <- alternates_census %>%
  filter(race!="AIAN" & race!="White" & race!="Black") %>%
  mutate(group=case_when(
    race %in% c("Chinese","Korean","Japanese","Vietnamese",
                     "Filipino","Cambodian","Hmong","Laotian",
                     "Thai","Burmese","Indonesian","Malaysian") ~ "E&SE Asian",
    race %in% c("Asian Indian","Bangladeshi",
                     "Pakistani","Sri Lankan") ~ "South Asian",
    race %in% c("Mexican","Puerto Rican","Cuban","Costa Rican",
                "Guatemalan","Honduran","Nicaraguan","Panamanian",
                "Salvadorian","Argentinian","Bolivian","Chilean",
                "Colombian","Ecuadorian","Paraguayan","Peruvian",
                "Uruguayan","Venezuelan","Dominican") ~ "Latino"),
    group=factor(group, levels=c("E&SE Asian","South Asian","Latino")))

diversity_ethnic <- temp %>%
  count(race, group, bpld) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980")

diversity_ethnic <- temp %>%
  count(race, group, lang) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980") %>%
  full_join(diversity_ethnic)

diversity_pan <- temp %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980")

diversity_pan <- temp %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="Census 1980") %>%
  full_join(diversity_pan)

temp <- alternates_acs %>%
  filter(race!="AIAN" & race!="White" & race!="Black") %>%
  mutate(group=case_when(
    race %in% c("Chinese","Korean","Japanese","Vietnamese",
                     "Filipino","Cambodian","Hmong","Laotian",
                     "Thai","Burmese","Indonesian","Malaysian") ~ "E&SE Asian",
    race %in% c("Asian Indian","Bangladeshi",
                     "Pakistani","Sri Lankan") ~ "South Asian",
    race %in% c("Mexican","Puerto Rican","Cuban","Costa Rican",
                "Guatemalan","Honduran","Nicaraguan","Panamanian",
                "Salvadorian","Argentinian","Bolivian","Chilean",
                "Colombian","Ecuadorian","Paraguayan","Peruvian",
                "Uruguayan","Venezuelan","Dominican") ~ "Latino"),
    group=factor(group, levels=c("E&SE Asian","South Asian","Latino")))

temp2 <- temp %>%
  count(race, group, bpld) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18")

diversity_ethnic <- temp %>%
  count(race, group, lang) %>%
  group_by(race, group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18") %>%
  full_join(temp2) %>%
  bind_rows(diversity_ethnic) %>%
  select(race, group, data, diversity_bpl, diversity_lang)

temp2 <- temp %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\nall ethnic groups")

diversity_pan <- temp %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\nall ethnic groups") %>%
  full_join(temp2) %>%
  bind_rows(diversity_pan)

races1980 <- as.character(subset(diversity_ethnic, data=="Census 1980")$race)

temp2 <- temp %>%
  filter(race %in% races1980) %>%
  count(group, bpld) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_bpl=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups")

diversity_pan <- temp %>%
  filter(race %in% races1980) %>%
  count(group, lang) %>%
  group_by(group) %>%
  mutate(p=prop.table(n)) %>%
  summarize(diversity_lang=1-sum(p^2)) %>% 
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups") %>%
  full_join(temp2) %>%
  bind_rows(diversity_pan)  %>%
  mutate(type="pan-ethnic group diversity")

temp2 <- diversity_ethnic %>%
  group_by(group, data) %>%
  summarize(diversity_bpl=mean(diversity_bpl),
            diversity_lang=mean(diversity_lang)) %>%
  ungroup() %>%
  mutate(data=ifelse(data=="ACS 2014-18", "ACS 2014-18\nall ethnic groups",
                     data))

diversity_pan <- diversity_ethnic %>%
  filter(race %in% races1980 & data=="ACS 2014-18") %>%
  group_by(group, data) %>%
  summarize(diversity_bpl=mean(diversity_bpl),
            diversity_lang=mean(diversity_lang)) %>%
  ungroup() %>%
  mutate(data="ACS 2014-18\n1980 ethnic groups") %>%
  bind_rows(temp2)  %>%
  mutate(type="mean of within ethnic group diversity") %>%
  bind_rows(diversity_pan) %>%
  pivot_longer(cols=starts_with("diversity_"), names_to="variable",
               values_to="diversity") %>%
  mutate(variable=ifelse(variable=="diversity_bpl",
                     "Birthplace",
                     "Language"),
         diversity=ifelse(group=="South Asian" & 
                            data!="ACS 2014-18\nall ethnic groups",
                          NA, diversity),
         data=factor(data,
                     levels=c("Census 1980", 
                              "ACS 2014-18\n1980 ethnic groups",
                              "ACS 2014-18\nall ethnic groups")),
         group=factor(group, 
                      levels=c("E&SE Asian", "Latino","South Asian")))
```

# Introduction

Social scientists have long treated patterns of intermarriage as a prime indicator of the underlying social distance and strength of boundaries separating groups. Because marriage is one of the most intimate kinds of social relationships, a widespread willingness to cross group boundaries is seen as a key signifier that the salience of those boundaries is weak [@gordon_assimilation_1964].

An increase in intermarriage across one boundary is often treated as an indication that this boundary is dissolving or blurring, but it can also indicate a reshaping of existing boundaries. Widespread intermarriage among European ethnic groups in the mid-twentieth century contributed to the breakdown of salient ethnic divisions between these populations but also helped to re-consolidate a sense of collective whiteness among these groups [@lieberson_many_1988; @alba_ethnic_1990; @jacobsen_whiteness_1998]. Similarly, spurred by resurgent immigration to the US since the 1960s, intermarriage is seen as a key benchmark for gauging panethnic affinity among Asian and Latino ethnic groups today.^[I use the term "ethnic groups" here to distinguish primarily different nationality groups among the Asian and Latino population, such as Chinese, Korean, Mexican, and Colombian.]

Panethnicity is defined by @okamoto_panethnicity_2014a as "the construction of a new categorical boundary through the consolidation of ethnic, tribal, religious, or national groups." Within the larger theoretical tradition studying ethnoracial boundary formation, panethnicity is treated as a form of boundary expansion in which the salient boundary shifts from one level to another within a nested hierarchy of possible identification [@wimmer_making_2008]. From this boundary perspective, ethnoracial categories are not fixed, stable, or primordial, but can shift over time in response to historical events, as they have for the ethnoracial groups in the US today that we collectively view as White and Black. Understanding how such processes are playing out for contemporary Asian and Latino ethnic groups informs us about how social boundaries might shift in the future, which has consequences for issues of racial inequality, assimilation, political change, and even how we measure race and ethnicity.

The earliest work on panethnicity focused on collaboration among ethnically based organizations and movements [@lopez_panethnicity_1990; @okamoto_theory_2003]. However, these authors also noted the importance of future work looking at patterns of marriage to better understand "interpersonal" panethnicity [@espiritu_asian_1993, pp. 167-168]. Some of the early attempts to address this research area simply looked at outmarriage percentages and thus incorrectly conflated group size difference with affinity [@shinagawa_asian_1996]. However, shortly after the turn of the century, several more sophisticated studies were published that found generally found stronger evidence of panethnicity among Asians than Latinos [ @qian_asian_2001; @rosenfeld_salience_2001; @qian_latinos_2004;  @fu_how_2007a;  @qian_crossing_2012]. 

This prior work on intermarriage has contributed greatly to our understanding of the phenomenon of panethnicity, but is is also considerably out of date and suffers from data limitations that have hampered its generalizability. Prior work has used data from Census 1980, 1990, or 2000. The two later data sources have no information on marriage timing which has made the identification and exclusion of immigrants married abroad from analysis difficult. Furthermore, much of the early work relies upon an examination of panethnicity among the relatively few Asian and Latino groups that were numerically large enough in historical data to sustain an analysis. Immigration from Asia and Latin America has diversified increasingly since this time  (CITATIONS?)

Additionally, much of the prior work on panethnic intermarriage is framed around issues of assimilation and acculturation. Researchers focus on differences in panethnic tendencies among the native-born and foreign-born. I argue instead that we should think about these issues from a larger social boundary framework, and that panethnicity should be considered a distinct process of boundary expansion. In the current demographic environment, issues of accculturation affect the tendency toward panethnicity but should be thought of as distinct and separable processes, rather than integral components of panethnicity. Specifically, individuals tend to prefer to marry someone born (and raised) in the same place and who speaks the same language. Birthplace and language endogamy have a complicated relationship to the tendency toward panethnic marriage. Our goal should be to estimate the likelihood of ethnic exogamy, accounting for these other tendencies rather than to collapse them all into an undifferentiated process of assimiliation/acculturation. Doing so allows us to see how much panethnic affinity there would be between panethnic groups in the absence of differences in birthplace and language thar are largely driven by continuing immigration from abroad.

In this article, I refresh our understanding of panethnic intermarriage using recent data from the 2014-2018 American Community Survey. I compare this data to Census 1980 data to analyze trends in panethnic intermarriage over time. Both of these data sources include timing of marriage which resolves several methodological difficulties with prior work. Additionally, I use a relatively novel counterfactual marriage model technique [@gullickson_counterfactual_2021] that allows me to easily adjust for differences in the spatial distribution of ethnic groups and to account for the role of birthplace and language endogamy when estimating the strength of panethnicity.

I find that the patterns of panethnicity are very different for Asian and Latino ethnic groups. For Asians, panethnic affinity is very strong among East/Southeast Asian ethnic groups and among South Asian ethnic groups, but there is little evidence of panethnicity between South Asian and East/Southeast Asian ethnic groups. Furthermore, this panethnic affinity has grown substantially over time. For Latinos, panethnic affinity is lower overall and in some cases is less likely than outmarraige to non-Latino racial groups. Additionally, panethnic affinity is reduced substantially in actual practice by significant language and birthplace differences for Asian ethnic groups, but not for Latinos. These different patterns have important implications for the future salience of panethnic categories.

# Pan-Ethnicity in Marriage

Panethnic affinity may be driven by both cultural and structural factors [@lopez_panethnicity_1990]. Cultural factors such as shared language and religion may create affinity across ethnic or national boundaries. Structural factors such as economic and occupational similarity, spatial proximity (through either segregation or settlement patterns), and the degree of racialization (the tendency of natives to treat all members of the panethnic category as a monolithic racial group) may also play a role in encouraging or discouraging panethnic affinity. The distinction between cultural factors and racializiation has been particularly important in differentiating the experiences of Asians and Latinos. Latinos share culture features such as language and religion that should promote panethnicity while Asians generally do not. On the other hand, racialization of Asians as a singular group tends to be stronger than for Latinos, who are more phenotypically diverse and racially ambiguous [@lopez_panethnicity_1990; @qian_latinos_2004]. 

Marriage patterns among members of the same panethnic group are seen as a key indicator of "on the ground" interpersonal panethnicity. Several important studies were performed shortly after the turn of the century on this topic, using data from the 1980 and 1990 US Census [@qian_asian_2001; @rosenfeld_salience_2001; @qian_latinos_2004; @fu_how_2007a]. Both @rosenfeld_salience_2001 and @fu_how_2007a explicitly compare the tendency toward panethnicity among Asians and Latinos and find a stronger tendency toward panethnicity among Asians. @qian_asian_2001 and @qian_latinos_2004 use similar methods to examine pan-ethnicity among Asians and Latinos, respectively, and find substantial evidence of panethnic affinity, particularly among the native-born. [@qian_crossing_2012] use 2000 Census data to estimate the likelihood of panethnic intermarriage for Mexicans, Puerto Ricans, Chinese, and Filipinos. Their results generally show similar levels of panethnicity across all four groups, and that panethnicity is most likely among the native-born and those immigrants who came to the US at an early age.

Some of these studies also compare how common interethnic marriage is in relation to interracial marriage with Whites and Blacks [@qian_asian_2001; @fu_how_2007a; @qian_crossing_2012]. In general, scholars have found that Asians are more likely to marry interethnically than to marry Whites or Blacks.  Latinos are also much less likely to marry Blacks than to marry interethnically, but in some cases, depending on generation and the ethnic group under consideration, may be somewhat more likely to marry Whites than to marry interethnically.

Although this prior work has enriched our understanding of panethnic intermarriage, it has also suffered from well-acknowledged methodological difficulties. All of the studies cited above use log-linear models to adjust for differences in group size. This approach allows scholars to properly estimate the underlying affinity between groups in the marriage market, apart from issues of exposure. However, basic log-linear models will not address differences in spatial patterns among groups. Most individuals will find a partner within a local, rather than a national, marriage market, and ethnoracial groups are not evenly distributed across the US. Without adjusting for these differences in geography, scholars will tend to underestimate the degree of ethnoracial exogamy [@harris_how_2005]. This is particularly important when studying panethnicity, because Latino and Asian ethnic groups have historically settled in different parts of the US (e.g. Cubans in Florida, Puerto Ricans in the Northeast, and Mexicans in the Southwest). Some of the prior work on panethnicity has attempted to address this issue. @fu_how_2007a includes Census division as a parameter in log-linear models, while @rosenfeld_salience_2001 estimated log-linear models separately within a few specific metropolitan areas. However, most of the prior work makes no adjustment for this important issue. 

All work on intermarriage has to wrestle with the difference between measuring the prevalence and incidence of intermarriage. Ideally, scholars wish to capture the incidence of intermarriage by examining recently formed marriages. However, most large-scale data has lacked information on marriage timing and so scholars have had to resort to restricting the age of spouses (e.g. limiting the analysis to marriages in which the wife is 20-34 years of age) in order to create semi-incidence measures with unknown biases. 

When addressing intermarriage among Latinos and Asians, this limitation is compounded by the issue of immigrants married abroad (IMA). Without information on both the timing of marriage and the timing of immigration to the US, it is impossible to completely remove IMA from the analysis. Because IMA were mostly married in their country of origin, they will bias estimates toward ethnic endogamy [@hwang_problem_1990]. Researchers have used a variety of sample restrictions in an attempt to minimize this problem, but without information on marriage and migration timing, it cannot be fully eliminated.

A final issue in estimating pan-ethnicity is the consideration of what ethnic groups to include. Historically, the most populous Latino ethnic groups were Mexicans, Puerto Ricans, and Cubans and for Asian ethnic groups, Chinese, Filipino, Japanese, Korean, and Vietnamese. Prior work have typically focused on some subset of these groups because they are the numerically largest and because they can be clearly identified in earlier Census data. However, immigration has increased substantially from areas in Asia and Latin America outside of these traditional sending areas and many more groups can be identified in Census data now. The role of increasing diversification of Asian and Latino population in panethnic intermarriage is an important consideration.

Related to this issue, most prior work has not addressed the issue of South Asian panethnic intermarriage. The placement of South Asians within the panethnic category of Asian is contested at best. Individuals from the Indian subcontinent are treated as part of the panethnic Asian group by governmental agencies such as the Census Bureau, but research has shown that in everyday practice they are more generally treated as "ambiguous non-whites" [@kibria_not_1996; @morning_racial_2001; @schachter_finding_2014]. This may be in part because the "somatic norm image" of what an Asian in the US contexts looks like does not correspond to South Asian populations [@kibria_not_1996]. However only @qian_asian_2001 have analyzed intermarriage data that explicitly includes South Asians. They found that Asian Indians had the lowest rate of intermarriage with other Asian groups, who were all East/Southeast Asian. 
    
## The Role of Birthplace and Language Endogamy

Research on intermarriage is often framed around issues of assimilation and acculturation. Historically, intermarriage was seen as a key benchmark for assimiliation into the "mainstream" culture [@gordon_assimilation_1964]. More recent work has problematized the assumption of one-way assimilation into the mainstream, most notably the work on segmented assimilation which suggests multiple pathways to assimilation for contemporary migrants [@portes_new_1993]. 

Panethnicity has sometimes been framed as a pathway within the segmented assimilation approach, specifically as a form of "selective acculturation" [@bohra-mishra_intermarriage_2015]. However, this framing seems somewhat forced as the original idea of selective acculturation was that an ethnic group would engage in social closure to facilitate resource sharing and the intergenerational transfer of cultural meaning and identity. Panethnic intermarriage would work against such social closure. 

The boundary approach to panethnicity utilized more broadly by many scholars would suggest that panethnicity is not simply a type of assimilation pattern, but rather a specific process of ethnoracial boundary expansion (CITATION - EXPAND ON THIS MORE). This boundary expansion is not simply a function of assimilation and acculturation but can be deeply affected by these cross-cutting issues. 

Most notably, birthplace and language endogamy will present challenges to the likelihood of panethnic intermarriage in complex ways. Individuals will often prefer partners from the same birthplace due to the shared cultural understandings that arise from being born and raised in a particular place. In general, we would expect birthplace endogamy to lower the likelihood of panethnicity for both Asians and Latinos because both of these groups have a substantial foreign-born population.

Similarly, people are more likely to marry individuals who speak the same primary language because of ease of communication. This form of endogamy is expected to affect Asians and Latinos differently, however. Language differences across Asian ethnic groups will lower the odds of panethnicity. Because Latinos share a common language, language endogamy should not serve as a significant barrier to Latino panethnicity.

However, the actual situation for both birthplace and language endogamy is considerably more complex than either of these summaries imply because ethnic groups themselves are also diverse in terms of birthplace and language. For example, among adults recently married or single in the 2014-18 American Community Survey date detailed below, 66% of Japanese respondents spoke English as a primary language, while 41% of Filipino respondents spoke English as a primary language. For those Japanese and Filipino individuals who both speak English, language endogamy will encourage panethnic intermarriage and discourage ethnic endogamy with members of their own group who speak Japanese and Tagalog, respectively. Similarly, 31% of Mexican Americans in this sample spoke English as primary language, but only 13% of Dominican Americans. This contrast is further complicated however by the fact that among Mexicans and Dominicans who do not speak English as a primary language, they will almost certainly speak Spanish as their primary language.  

This diversity in language and birthplace *within* ethnic groups makes the relationship of birthplace and language endogamy to panethnic intermarriage complex. In order for language and birthplace endogamy to serve as a barrier to panethnic intermarriage, the diversity in language and birthplace endogamy must be greater within the panethnic group than within the corresponding ethnic group. The strength of the barrier will depend on how much more diverse the panethnic group is than the ethnic group. These patterns are complex and have not been fully appreciated by prior work on the subject, which often addresses these issues simply by dividing the sample between the immigrant and native-born population. 

Ideally, to estimate the underlying affinity for panethnic intermarriage, we want a model that can separate out those barriers to intermarriage resulting from language and birthplace endogamy. Such a model would give us a counterfactual case of how much panethnicity we would expect in a situation in which all Latinos and Asians are fully assimilated/acculturated to the US (i.e. born in the US and speak English as their primary language).

In the analysis that follows, I address many of the weaknesses with prior data and models using an  counterfactual marriage model to estimate the likelihood of panethnic intermarriage. I use newly available American Community Survey data from 2014-18 that includes the timing of the current marriage and I compare this to 1980 Census data that also contains information on the timing of marriage. Using this marriage timing, I am able to exclude immigrants born abroad from the analysis and generate a true incidence measure of panethnicity that captures recent trends. I use this model to compare to trends in panethnic intermarriage for Latinos, East/Southeast Asians, and South Asians.

# Data and Methods

The data for this analysis are derived from the microdata sample of the 1980 US Census and the American Community Survey (ACS) for 2014-2018. The ACS is an annual 1-in-100 survey of the United States population, conducted by the United States Census Bureau. The 1980 US Census is the last Census to include information about marriage timing. Information about marriage timing was added to the ACS in 2008. To increase sample size for smaller ethnoracial categories, I pool five years of ACS data from 2014-2018. Both data sources were extracted from the Integrated Public Use Microdata Series (IPUMS) system [@ruggles_ipums_2020].

I use a counterfactual marriage model [@qian_marriage_2018; @gullickson_counterfactual_2021] for the analysis that follows. In both data sources, I restrict the analysis to all opposite-sex marriages formed in the previous five years that were a first marriage for both partners. The restriction to opposite-sex first marriages is necessary for comparison because the Census 1980 only recorded marriage timing relative to first marriage and does not include same-sex unions. For each marriage, I then construct a choice set of one real union and twenty fictional unions. Fictional unions are created by sampling alternate partners for one randomly determined spouse from a pool of potential partners. I then use a conditional logit model to predict how partner characteristics relate to the likelihood of observing the true union, as follows: $$P_{ij}=\frac{e^{\mathbf{x}_{ij}\beta}}{\sum_{k=1}^J e^{\mathbf{x}_{ik}\beta}}$$ where $P_{ij}$ is the probability that union $j$ within choice set $i$ is the actual union. $J$ is the total number of unions in the choice set (in this case, twenty-one). The vector $\mathbf{x}_{ij}$ defines the characteristics of the union and the $\beta$ vector provides estimated log-odds ratios indicating how the odds of an actual union change with $\mathbf{x}_{ij}$. The model is estimated as a fixed-effects logistic regression model with fixed effects for each choice set.

One advantage of the counterfactual marriage model is that I can easily set up parameters for sampling alternate partners that better capture the context of local marriage markets. The set of potential alternate partners from which fictional spouses are sampled is made up of all unmarried adults as well as those who were married in the previous five years. Alternate partners are sampled from among individuals residing in the same geographically defined marriage market who have not migrated to the US or across state lines in the last five years. I define marriage markets by the metropolitan area of residence when this can be identified for a respondent, and otherwise by the state of residence. Metropolitan area is not directly identifiable in the data used here due to confidentiality concerns, but can be deduced for many metropolitan areas based on the finest geographic identifier of the respondent (county group in Census 1980 and Public Use Microdata Area in the ACS). IPUMS provides metro area identification for respondents in cases where the county group or Public Use Microdata Area was mostly contained within the metropolitan area. Figure \@ref(fig:metro-areas) shows the identifiable metropolitan areas used in the analysis for both time periods. Although not all metropolitan areas are identified in the data, most major metropolitan areas are included in both time periods. Coverage differs somewhat by period, with more extensive coverage in the later time period. 

```{r metro-areas, fig.cap="Metropolitan statistical areas identifiable in the data by data source. Fill colors are drawn with partial transparency so that darker areas indicate overlapping areas in the two data sources."}
knitr::include_graphics(here("analysis","output","msa_map.pdf"), 
                        auto_pdf = FALSE)
```

Because this approach relies upon a random sample of alternate partners, results will vary each time this sampling procedure is performed. To account for the added uncertainty due to sampling, I perform the sampling procedure separately five times to produce five separate analytical datasets. I then run parallel models on all five datasets and pool $\beta$ estimates and standard errors using the same methods as those employed for multiple imputation [@rubin_multiple_1987; @gullickson_counterfactual_2021].

## Variable Construction

One advantage of the counterfactual marriage model is that it allows for the easy incorporation of a variety of quantitative and categorical variables. I utilize that feature here to account for age differences, educational differences, birthplace endogamy, and language endogamy when estimating the likelihood of panethnic intermarriage. 

I measure ethnoracial categories by a combination of the race and Hispanicity questions in both data sources. Table \@ref(tab:sample-size-table) shows the sample size for all ethnoracial groups included in the analysis. I classify non-Asian and non-Latino respondents as White, Black, or American Indian/Alaska Native (AIAN). Due to small sample size, I exclude Pacific Islanders, multiracial respondents, and non-Latino others from the analysis. The race question includes several Asian nationalities (e.g. Chinese, Korean, Filipino) as well as a write-in response for ethnic/national identities not captured by the existing categories. Similarly, the Hispanicity question includes major Latino ethnic groups (e.g. Mexican, Puerto Rican, Cuban) as well as a write-in option. Importantly, the Census 1980 data includes a far more limited set of identifiable ethnic groups for both Asians and Latinos than the ACS data. In Census 1980, I can only identify Mexicans, Puerto Ricans, and Cubans among Latinos and Chinese, Japanese, Korean, Filipino, Vietnamese, and Asian Indian respondents among Asians. To address this restriction, I estimate two different kinds of models for the ACS data. In the first model, I restrict the data only to those ethnic groups that were available in the Census 1980 data to allow more direct comparison over time. I also estimate models using the full ACS data to explore how using a broader set of ethnic groups changes our understanding of panethnicity, but these models are not directly comparable to the Census 1980 models.

```{r sample-size-table}

temp <- c(table(alternates_acs$race)) %>% enframe()
tab_samp <- c(table(alternates_census$race)) %>% 
  enframe() %>%
  right_join(temp, by="name") %>%
  rename(ethnorace=name, census=value.x, acs=value.y) %>%
  mutate(race_big=case_when(
    ethnorace=="Chinese" ~ "East and Southeast Asian",
    ethnorace=="Japanese" ~ "East and Southeast Asian",
    ethnorace=="Korean" ~ "East and Southeast Asian",
    ethnorace=="Filipino" ~ "East and Southeast Asian",
    ethnorace=="Vietnamese" ~ "East and Southeast Asian",
    ethnorace=="Cambodian" ~ "East and Southeast Asian",
    ethnorace=="Hmong" ~ "East and Southeast Asian",
    ethnorace=="Laotian" ~ "East and Southeast Asian",
    ethnorace=="Thai" ~ "East and Southeast Asian",
    ethnorace=="Burmese" ~ "East and Southeast Asian",
    ethnorace=="Indonesian" ~ "East and Southeast Asian",
    ethnorace=="Malaysian" ~ "East and Southeast Asian",
    ethnorace=="Bhutanese" ~ "East and Southeast Asian",
    ethnorace=="Nepalese" ~ "East and Southeast Asian",
    ethnorace=="Asian Indian"  ~ "South Asian",
    ethnorace=="Pakistani" ~ "South Asian",
    ethnorace=="Bangladeshi" ~ "South Asian",
    ethnorace=="Sri Lankan" ~ "South Asian",
    ethnorace=="Mexican" ~ "Latino", 
    ethnorace=="Cuban" ~ "Latino", 
    ethnorace=="Puerto Rican" ~ "Latino", 
    ethnorace=="Guatemalan" ~ "Latino", 
    ethnorace=="Salvadorian" ~ "Latino", 
    ethnorace=="Dominican" ~ "Latino", 
    ethnorace=="Colombian" ~ "Latino", 
    ethnorace=="Costa Rican" ~ "Latino", 
    ethnorace=="Honduran" ~ "Latino", 
    ethnorace=="Nicaraguan" ~ "Latino", 
    ethnorace=="Panamanian" ~ "Latino", 
    ethnorace=="Argentinian" ~ "Latino", 
    ethnorace=="Bolivian" ~ "Latino", 
    ethnorace=="Chilean" ~ "Latino", 
    ethnorace=="Ecuadorian" ~ "Latino", 
    ethnorace=="Peruvian" ~ "Latino", 
    ethnorace=="Venezuelan" ~ "Latino", 
    ethnorace=="Paraguayan" ~ "Latino", 
    ethnorace=="Uruguayan"  ~ "Latino"
  ))

temp <- tab_samp %>% group_by(race_big) %>%
  summarize(census=sum(census, na.rm=TRUE), acs=sum(acs, na.rm=TRUE)) %>%
  na.omit() %>%
  rename(ethnorace=race_big)

tab_samp <- tab_samp %>% select(ethnorace, census, acs) %>%
  bind_rows(temp)

temp <- tibble(ethnorace="Marriages in the previous five years",
               census=census_n, acs=acs_n)

tab_samp <- tab_samp %>% 
  bind_rows(temp)
                    
ord <- c("Marriages in the previous year",
         "White","Black","AIAN","Latino",
         "Mexican","Puerto Rican","Cuban","Salvadorian","Dominican","Guatemalan",
         "Colombian","Honduran","Peruvian","Ecuadorian","Nicaraguan","Argentinian",
         "Venezuelan","Panamanian","Chilean","Costa Rican","Bolivian",
         "Uruguayan","Paraguayan",
         "East and Southeast Asian",
         "Chinese","Filipino","Vietnamese","Korean","Japanese","Cambodian",
         "Hmong","Laotian","Thai","Burmese","Indonesian","Malaysian",
         "South Asian",
         "Asian Indian","Pakistani","Bangladeshi","Sri Lankan")

tab_samp <- as.data.frame(tab_samp)
rownames(tab_samp) <- tab_samp$ethnorace
tab_samp <- tab_samp[ord,]
rownames(tab_samp) <- NULL

options(knitr.kable.NA = '')

tab_samp %>%
  mutate(census=format(census, big.mark=","),
         census=ifelse(census=="       NA","",census),
         acs=format(acs, big.mark=",")) %>%
  kbl(caption = "Sample size of marriages and alternate partners by data source and ethnoracial category.", booktabs=T, longtable=F,
      linesep = NULL,
      col.names=c("Category","Census 1980", "ACS 2014-2018"),
      align = c("l","r","r")) %>%
  kable_styling(full_width = TRUE) %>%
  add_indent(c(6:24,26:37,39:42)) %>%
  pack_rows("Marriages", 1, 1) %>%
  pack_rows("Alternate Partners", 2, 42) %>%
  column_spec(1, width = "6cm")
```

I model interracial and panethnic marriage using a set of dummy variables where the reference category is a racially and ethnically endogamous union (e.g. a White/White or Chinese/Chinese marriage). Because the number of ethnoracial categories is large, modeling every possible combination of these categories for exogamous unions is not feasible. In my most restrictive models, I apply two limits on the parameterization that make model estimation possible. First, when analyzing intermarriage with Whites, Blacks, and American Indians/Alaska Natives, I group Asian and Latino ethnic groups into the larger panethnic categories of Latino, East/Southeast Asian, and South Asian. For example, a union involving a Black and Mexican partner and a union involving a Black and Colombian partner would both be coded as a Black/Latino union. Second, I model the likelihood of panethnic intermarriage using a single dummy variable to indicate any form of ethnic exogamy within the three panethnic categories of Latino, East/Southeast Asian, and South Asian. For example, a union involving a Chinese and Japanese partner and a union involving a Korean and Filipino partner would both be coded as East/Southeast Asian ethnic exogamy. This approach allows me to estimate the average likelihood of ethnic exogamy across all possible combinations, but does not capture particular affinities of disaffinities among ethnic groups within the panethnic category. I partially relax both of these these assumptions in later models of the ACS data on a sample restricted to the largest Asian and Latino ethnic groups in order to estimate affinities specific to ethnic groups.

Because I separate South Asian from East/Southeast Asian groups and only Asian Indians are identifiable in the Census 1980 data, I cannot estimate panethnic parameters over time for South Asians. Thus, I focus primarily on panethnic parameters among East/Southeast Asians and Latinos in the analysis that follows. I do estimate and report panethnic parameters for South Asians for some models using ACS data and I include a parameter for intermarriage between East/Southeast Asians and South Asians in all models. 

I account for language and birthplace endogamy with simple dummy variables indicating whether the two potential partners share the same primary language or birthplace, respectively. Primary language is determined by what language the respondent reported speaking at home. Because this variable is measured after a marriage has occurred, it may somewhat overestimate language endogamy among respondents. Some partners who spoke different languages when they met may have shifted the language spoken at home after marriage.

Birthplace endogamy is complicated by the "1.5" generation -- individuals who were born outside of the United States but migrated as children and whose formative experiences are at least partially defined by acculturation within the US. Within the coding scheme used here, I consider three possibilities for such individuals, organized by the assumed degree of acculturation and assimilation. First, such individuals could be considered birthplace endogamous only with people from the same actual birthplace. Second, such individuals could be considered birthplace endogamous with either a person from their actual birthplace or the a person born in the USA. Third, such individuals could be considered only birthplace endogamous with people born in the USA.

To test the accuracy of these three scenarios, I fit models using each coding scheme to the Census 1980 and full ACS data. Following @rumbaut_ages_2004a, I also divide the 1.5 generation into a "1.75" generation (those who arrived in the US before the age of 6), a "1.5" generation (those who arrived in the US between the ages of 6-12), and a "1.25" generation (those who arrived in the US between the ages of 13-17). For these three groups, I consider every possible combination of coding such that earlier generations are not more acculturated than later generations (e.g. the 1.25 generation could not be "both" in cases where the 1.5 generation is "birthplace only"). Table \@ref(tab:deviance-bendog) below shows the model fit of all ten possible combinations for both data sources.

```{r deviance-bendog}
dev_acs <- sapply(models_bendog_acs, function(x) {mean(x$deviance)}) %>%
  enframe(name="model", value="deviance_acs")

sapply(models_bendog_census, function(x) {mean(x$deviance)}) %>%
  enframe(name="model", value="deviance_census") %>%
  left_join(dev_acs) %>%
  mutate(gen1.75=c("Birthplace","USA","Both","USA","USA","USA","USA", "USA", 
                   "Both","Both"),
         gen1.5 =c("Birthplace","USA","Both","Both","Birthplace","USA","Both",
                   "USA", "Birthplace","Both"),
         gen1.25=c("Birthplace","USA","Both","Birthplace","Birthplace",
                   "Birthplace","Both","Both","Birthplace","Birthplace"),
         deviance_census=format(deviance_census, big.mark=","),
         deviance_acs=format(deviance_acs, big.mark=","),
         deviance_census=cell_spec(deviance_census,
                                   bold = ifelse(deviance_census=="1,122,736",
                                                       TRUE, FALSE)),
         deviance_acs=cell_spec(deviance_acs,
                                   bold = ifelse(deviance_acs=="1,849,202",
                                                       TRUE, FALSE)),
         ) %>%
  select(gen1.25, gen1.5, gen1.75, deviance_census, deviance_acs) %>%
  relocate(gen1.75, gen1.5, gen1.25, deviance_census, deviance_acs) %>%
  arrange(gen1.25, gen1.5, gen1.75) %>%
  kbl(caption = "Model fit to Census 1980 and ACS data using different specifications of birthplace endogamy for 1.25, 1.5, and 1.75 generations. Minimum deviance is shown in bold.", booktabs=T, longtable=F, 
      col.names=c("1.75","1.5","1.25","Census 1980","ACS 2014-18"),
      align = c("l","l","l","r","r"),
      linesep = NULL, escape = FALSE) %>%
  kable_styling(full_width = TRUE) %>%
  add_header_above(c("Generation" = 3, "Model Deviance" = 2)) %>%
  #column_spec(4, background = "red") %>%
  #column_spec(1, width = "5cm") %>%
# Goddmmit: apparently I need to use longtable to get wrapped footnotes, but then it won't float
  footnote(general=c("Age of US arrival by generation is 0-5 (1.75), 6-12 (1.5), and 13-17 (1.25). All models include",
                     "controls for educational differences, age differences, ethnoracial endogamy, and language",
                     "endogamy."), 
        threeparttable=F, footnote_as_chunk = F, 
          general_title="Notes: ", title_format="italic")
```

For both time periods, the most preferred models use birthplace only coding for those respondents who entered the US after age 12. The USA only option representing full acculturation was not preferred for any group in either time period. In the Census 1980 data, the most preferred model treated all three groups as first generation individuals and only considered them endogamous with partners from the same actual birthplace. The more recent ACS data preferred a model in which both the 1.75 and 1.5 generations were considered birthplace endogamous both with partners from the actual birthplace or the USA. The shift over time suggests greater acculturation of the 1.5 and 1.75 generation today than in 1980. For all models in the analysis that follows, I code birthplace endogamy according to the best-fitting model for each data source from Table \@ref(tab:deviance-bendog).

All models used here include controls for age and educational differences between potential partners. Age differences are modeled by taking the numerical age difference between partners and its square. Educational differences are modeled using educational crossing parameters [@schwartz_trends_2005]. All respondents are categorized as having less than a high school diploma, a high school diploma, some college education, or at least a four-year college degree. Three dummy variables indicate whether a particular union involves crossing the adjacent pairwise boundaries between these four categories. Results are cumulative such that a marriage involving one partner with less than a high school education and one partner with a four-year college degree would require passing all three boundaries.

In the results below, I use graphical visualizations to present important results from the models. In all cases, results are presented as the ratio of the odds of a particular kind of exogamy (racial or ethnic) to the odds of ethnoracial endogamy. An odds ratio of one indicates no barrier to a given kind of exogamy. The full model results upon which figures are based are available in tabular form in the supplementary materials for this article. 

# Results

## Birthplace and Language Diversity

Before moving to the model results, I first show the complex ways in which birthplace and language endogamy affect the likelihood of panethnic intermarriage. I use the Blau index [@blau_inequality_1977] to measure language/birthplace diversity within each Asian and Latino ethnic group and to measure diversity within each panethnic group. The Blau index of diversity ($D$) is calculated as: $$D=1-\sum_{i=1}^R p_i^2$$ where $p_i$ is the proportion of the total population belonging to group $i$ and $R$ is the total number of groups. The Blau index has an intuitively appealing interpretation as the probability that two randomly drawn members from the population belong to the same group.

```{r diversity-pan-bar, fig.cap="Birthplace and language diversity within Asian and Latino ethnic groups compared to birthplace and language diversity within pan-ethnic groups. Diversity is measured by the Blau index which gives the probability that two randomly selected members of the group do not share the same birthplace/language."}
ggplot(diversity_pan, aes(y=group, x=diversity, color=type))+
  #geom_col(position="dodge")+
  geom_vline(xintercept = 0, color="grey20", linetype=1)+
  geom_pointrangeh(aes(xmin=0, xmax=diversity), size=0.4,
                   position=position_dodgev(height=0.6))+
  facet_grid(variable~data)+
  scale_x_continuous(labels = scales::percent, breaks = c(0, 0.25, 0.5, 0.75))+
  scale_color_manual(values = my_palette[1:2])+
  labs(y=NULL, x="Probability two members do not share the same birthplace/language",
       fill="type of diversity")+
  theme_mine()+
  theme(panel.grid.major.y = element_blank())
```

Figure \@ref(fig:diversity-pan-bar) shows the Blau index for language and birthplace across East/Southeast Asian, South Asian, and Latino ethnic and panethnic groups, using data on alternate partners in the two time periods. To simplify presentation, I take the mean diversity within each specific ethnic group and compare this directly to the diversity across the panethnic group as a whole. Barriers to panethnicity will be higher in cases where panethnic diversity is higher than average diversity within ethnic groups.

Significant language and birthplace diversity is observable within specific ethnic groups (e.g. Chinese, Mexican). For example, the average diversity in birthplace within specific Latino ethnic groups in the ACS data is nearly 60%, indicating that two randomly determined members of the same ethnic group would not share the same birthplace about 60% of the time. This diversity is mostly driven by within ethnic group differences between being born in the place of origin or being born in the US and speaking the common language of the origin group or speaking English. Panethnic diversity on the other hand is driven by additional differences in non-US birthplace and non-English language across specific ethnic groups.

Figure \@ref(fig:diversity-pan-bar) shows that panethnic diversity in language and birthplace is greater than within ethnic group diversity for all three panethnic groups in both time periods. However, these differences are small for Latinos, while they are quite large for Asians. Because Latinos share a common non-English language, this result was expected for language, but I also observe a similar pattern here for birthplace. What differences exist among Latino ethnic groups have also diminished over time, whereas they have grown for East/Southeast Asians.

The consequence of these patterns is that language and birthplace endogamy is an important barrier to panethnic intermarriage among Asians, but plays little role in panethnic intermarriage among Latinos. With that in mind, I now turn to models that estimate the likelihood of panethnic intermarriage with and without accounting for birthplace and language endogamy.

## Modeling Panethnicity

Figure \@ref(fig:exog-lolly) shows the estimated odds of ethnic exogamy relative to ethnic endogamy for all three panethnic groups in both time periods. As these odds ratios approach one, the odds of ethnic exogamy (e.g. Chinese person marrying a Japanese person) increase relative to ethnic endogamy (e.g. Japanese person marrying a Japanese person). Due to data limitations, results are only available for South Asians in the more recent ACS data.

```{r exog-lolly, fig.cap="The odds of ethnic exogamy for Latinos, East/Southeast Asians, and South Asians over time and model specification. Results are based on counterfactual marriage models. The baseline model controls for educational assortative mating and age differences."}
coefs_pent %>% 
  filter(variable=="E&SE Asian ethnic exogamy" | 
           variable=="Latino ethnic exogamy" |
           variable=="South Asian ethnic exogamy") %>%
  ggplot(aes(y=model, x=exp(coef), color=data))+
  geom_vline(xintercept = 0, color="grey20", linetype=1)+
  geom_vline(xintercept = 1, color="grey20", linetype=3)+
  geom_pointrangeh(aes(xmin=0, xmax=exp(coef)), size=0.4,
                   position=position_dodgev(height=0.6))+
  scale_color_manual(values=my_palette)+
  scale_x_continuous(limits = c(0,1))+
  labs(y=NULL, 
       x="Odds ratio of ethnic intermarriage relative to ethnic endogamy",
       color="Data source")+
  facet_wrap(~variable)+
  theme_mine()+
  theme(panel.spacing = unit(0.6, "lines"))
```

The baseline model includes only controls for age and educational differences between potential partners. I then include birthplace and language endogamy separately and together to determine the effects of these variables individually and collectively on the odds of ethnic exogamy.

In the baseline model, the odds of ethnic exogamy are similar for all three groups. In all cases, ethnic exogamy is about 25% as likely as ethnic endogamy. The results also show a very slight decline from the 1980 time period to the 2014-2018 when restricting to ethnic groups that were available in the 1980 data.

For Latinos, adjusting for birthplace and language endogamy has little effect on the odds ratio. Controlling for each variable individually increases the overall odds of ethnic exogamy only very slightly, with birthplace endogamy having a slightly larger effect. Overall, however, the odds are hardly changed at all. Furthermore, there is no evidence of change across the two time periods when limiting the analysis to comparable ethnic groups. In the later time period, ethnic exogamy is higher when including a larger range of ethnic groups, but still remains relatively low.

For both Asian panethnic groups, controlling for birthplace and language endogamy substantially increases the odds of ethnic exogamy, to the point that ethnic exogamy in the later time period is not at all unusual. For example, using the more expansive ethnic group coding, the odds of ethnic exogamy are roughly 75% as high as the odds of ethnic endogamy for both East/Southeast Asians and South Asians once I control for both language and birthplace endogamy. Controlling for language endogamy produces larger changes than controlling for birthplace endogamy, but both variables produce substantial changes.

Controlling for birthplace and language endogamy also reveals significant increases in the odds of ethnic exogamy over time for East/Southeast Asians. Using comparable ethnic groups, the odds of ethnic exogamy for East/Southeast Asians were about half those of ethnic endogamy in 1980, but have grown to 75% by the more recent time period. As for the Latino case, the odds of ethnic exogamy are slightly higher in the later time period when I consider all East/Southeast Asian ethnic groups rather than just those ethnic groups available in the 1980 data.

Figure \@ref(fig:exog-lolly) shows the important role that birthplace and language endogamy play in panethnic intermarriage. In actuality, we observe similar odds of ethnic exogamy among Asians and Latinos. However, the low odds of ethnic exogamy for Asians are largely a function of the high level of diversity in birthplace and language among Asians combined with a preference for marrying someone of the same birthplace and language. These barriers do not exist for Latinos. Controlling for these issues reveals very different affinities for panethnicity among Latinos and Asians in the absence of birthplace and language barriers.

```{r changes-intermar, fig.cap="Odds of ethnoracial exogamy relative to endogamy across two time periods. Results are based on models that control for age differences, educational differences, and birthplace and language endogamy. Values are sorted by ethnoracial exogamy in 1980. Arrows show the change across the two time periods based on comparable sets of ethnic groups. Results for American Indian/Alaska Native intermarriage are excluded due to sampling variability."}
coefs_pent %>%
  filter(model=="+ birthplace and\nlanguage endogamy" & 
           !grepl("South Asian ethnic exogamy", variable) & 
           !grepl("AIAN", variable)) %>%
  left_join(arrows_pent) %>%
  mutate(variable=str_replace(variable, "E&SE Asian ethnic exogamy", "**E&SE Asian ethnic exogamy**"),
         variable=str_replace(variable, "Latino ethnic exogamy", "**Latino ethnic exogamy**"),
         coef=ifelse(data=="ACS 2014-18, all ethnic groups" & variable=="White/Black", NA, coef)) %>%
  order_variables("Census 1980", "+ birthplace and\nlanguage endogamy") %>%
  ggplot(aes(x=variable, y=exp(coef)))+
  geom_hline(yintercept = 1, linetype=3, size=1)+
  geom_point(aes(color=data, shape=data), alpha=1, size=3)+
  geom_segment(aes(x=variable, xend=variable,
                   y=exp(coef.census), yend=exp(coef.acs)),
               color="grey30",
               arrow = arrow(length = unit(0.03, "npc")))+
  coord_flip()+
  scale_y_continuous(limits=c(0,1.1), expand=c(0.01,0))+
  scale_color_manual(values = my_palette)+
  scale_shape_manual(values = c(16,16,17))+
  labs(x=NULL, 
       y="Odds ratio of exogamy relative to ethnoracial endogamy",
       title=NULL,
       color="Data source", shape="Data source")+
  theme_mine()+
  theme(axis.text.y=element_markdown())
```

Figure \@ref(fig:changes-intermar) shows the odds of East/Southeast Asian and Latino ethnic exogamy in comparison to the odds of interracial marriage more broadly. Figure \@ref(fig:changes-intermar) also shows how these odds have changed over time. All results in Figure \@ref(fig:changes-intermar) are from the model that accounts for birthplace and language endogamy. East/Southeast Asian ethnic exogamy stands out as far more likely than any form of interracial marriage. Latino ethnic exogamy, on the other hand, is comparable in likelihood to several forms of interracial marriage. White/Latino intermarriage was less likely than Latino ethnic exogamy in 1980, but has increased in frequency substantially. The consequence of this shift is that White/Latino exogamy is slightly more likely than Latino ethnic exogamy in the more recent time period, although this is only true when looking at the Mexican, Puerto Rican, and Cuban ethnic groups identifiable in the 1980 data. When the analysis is expanded to all Latino ethnic groups, Latino ethnic exogamy is sightly more likely than Latino/White exogamy.

Figure \@ref(fig:changes-intermar) also shows the results for South Asian interracial marriage over time. The results show significant declines in South Asian interracial marriage with all other groups over the time period. These results, however, should probably be viewed with caution. In the earlier time period, the only South Asian group available was "Asian Indian" and research has shown that respondents sometimes confuse the categories of "Asian Indian" and "American Indian" in race responses [@liebler_american_2004]. Given the novel character of the "Asian Indian" category in 1980, it is likely that such misreporting has gone down over time. This may contribute to the overall decline in interracial marriage over time, if American Indians are less endogamous, on average, than Asian Indians.^[Although not shown here, I also found that the odds of intermarriage between American Indian/Alaska Natives and South Asians was roughly 1.9 to 1 in 1980 but declined to 0.103 to 1 in the later ACS data. This result suggests that such race reporting errors played a greater role in 1980 than the later data.] Nonetheless, the results here show a relative low odds of intermarriage between South Asians and East/Southeast Asians in both time periods, suggesting that Asian panethnicity occurs within two separate "melting pots" for South Asians and all other Asian groups.

Figure \@ref(fig:changes-intermar) also documents that the lowest likelihood of intermarriage in both time periods is that between Blacks and all other groups, despite some substantial increases in these odds. These results indicate the enduring strength of the Black/non-Black boundary in the US despite increasing diversity in the overall population.

## Group-Specific Affinities

The preceding models used a single ethnic exogamy term for each panethnic group such that the odds of ethnic exogamy were assumed to be the same regardless of which specific groups were paired. For example, the odds of ethnic exogamy into a Japanese-Korean union would be the same as those for a Japanese-Filipino union. I now turn to models of the ACS data that relax this assumption and allow the estimation of specific affinities and disaffinities within the matrix of ethnic groups that make up the three panethnic groups.

Fitting such models is actually quite difficult. If I considered every possible combination within panethnic groups in the ACS data, I would need to replace my three ethnic exogamy terms with 243 group specific parameters. Given that some of these groups are quite small in the dataset, models could not be fit to this fully expanded set, due to issues with sparseness and model convergence. I experimented with the largest possible set of ethnic groups for which I could reliably fit data and found that this model included the five East/Southeast Asian ethnic groups available in the Census 1980 data (Chinese, Filipino, Vietnamese, Korean, and Japanese) and the ten largest Latino groups, with the exception of Hondurans. I could not fit ethnicity specific models for South Asians because of the small size of non-Asian Indian groups in the South Asian category .

In addition to separate parameters for each specific pairing of ethnic exogamy, I also include ethnic-group specific parameters for intermarriage with Whites and Blacks. These parameters allow me to examination variation for each specific ethnic group in the tendency to intermarry with Whites and Blacks. 

Figure \@ref(fig:ethnic-heat-map) shows the group-specific odds of ethnic exogamy for the five East/Southeast Asian ethnic groups and the nine Latino ethnic groups, based on these models. I also provide a dendrogram that indicates the relative distance between each of these ethnic groups based on a tree diagram. The dendrogram is calculated by treating the inverse of the odds ratio as a measure of distance.


```{r ethnic-heat-map, fig.cap="Odds of ethnic exogamy relative to ethnic endogamy between particular E&SE Asian/Latino ethnic groups in ACS 2014-2018 data. The uppder panel shows E&SE Asian ethnic groups and the lower panel shows Latino ethnic groups. Dendrograms on the right are based on hierarchical clustering using unweighted average distances where distance was measured by theinverse of these odds ratios. Results are based on models that control for age differences, education differences, and language and birthplace endogamy.", fig.height=7}
asian_groups <- c("Chinese","Korean","Japanese","Vietnamese","Filipino")
or_asian <- coefs_extended %>%
  filter(data=="ACS 2014-18, expanded ethnic groups" &
           model=="+ both" &
           grepl(paste(asian_groups,collapse="|"), variable) &
           !grepl("Black|White", variable)) %>% #ugly hack, FIXME
  calc_or_matrix()
hc_asian <- hclust(as.dist(or_asian), method = "average")
or_asian <- or_asian[hc_asian$order, rev(hc_asian$order)]

gg.heat.asian <- ggplot(melt(1/or_asian, na.rm=TRUE), aes(Var2, Var1, fill = value))+
  geom_tile(color = NA)+
  scale_fill_gradient2(low = "#7b3294", high = "#008837", mid = "#f7f7f7", 
                       midpoint = 1, limit = c(0, 1.25),
                       name="odds ratio")+
  geom_text(aes(label=paste(round(1/or_asian,2))))+
  labs(x=NULL, y=NULL)+
  theme_minimal()+
  theme(legend.position = "left", 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1))

dendro_asian <- as.dendrogram(hc_asian)
gg.dendro.asian <- ggdendrogram(data = dendro_asian, rotate = TRUE, 
                                labels = FALSE)+
  theme_void()

latino_groups <- c("Mexican","Puerto Rican","Cuban","Guatemalan",
            "Salvadorian","Colombian","Ecuadorian","Peruvian",
            "Dominican")
or_latino <- coefs_extended %>%
  filter(data=="ACS 2014-18, expanded ethnic groups" &
           model=="+ both" &
           grepl(paste(latino_groups,collapse="|"), variable) &
           !grepl("Black|White", variable)) %>% #ugly hack, FIXME
  calc_or_matrix()
hc_latino <- hclust(as.dist(or_latino), method = "average")
or_latino <- or_latino[hc_latino$order, rev(hc_latino$order)]

gg.heat.latino <- ggplot(melt(1/or_latino, na.rm=TRUE), aes(Var2, Var1, fill = value))+
  geom_tile(color = NA)+
  scale_fill_gradient2(low = "#7b3294", high = "#008837", mid = "#f7f7f7", 
                       midpoint = 1, limit = c(0, 1.25),
                       name="odds ratio")+
  geom_text(aes(label=paste(round(1/or_latino,2))), size=2.5)+
  labs(x=NULL, y=NULL)+
  theme_minimal()+
  theme(legend.position = "left", 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 30, hjust = 1))

dendro_latino <- as.dendrogram(hc_latino)
gg.dendro.latino <- ggdendrogram(data = dendro_latino, rotate = TRUE, labels = FALSE)+
  theme_void()

grid.newpage()
print(gg.heat.asian, vp = viewport(x = 0.4, y = 0.75, width = 0.8, height = 0.5))
print(gg.dendro.asian, vp = viewport(x = 0.9, y = 0.79, width = 0.2, height = 0.34))
print(gg.heat.latino, vp = viewport(x = 0.4, y = 0.25, width = 0.8, height = 0.5))
print(gg.dendro.latino, vp = viewport(x = 0.9, y = 0.29, width = 0.2, height = 0.39))
```

The top panel of figure \@ref(fig:ethnic-heat-map) shows no social distance between the three East Asian groups of Chinese, Korean, and Japanese. The point estimates actually suggest that, once language and birthplace endogamy are accounted for, the odds of ethnic exogamy are slightly higher than ethnic endogamy among these groups. However, none of these odds ratios are statistically distinguishable from one using a p-value cutoff of 0.05. 

Social distance remains between these East Asian groups and the Vietnamese and Filipino ethnic groups, with the exception of the Vietnamese/Chinese case for which there is no barrier to ethnic exogamy. Filipinos, in particular, have the lowest odds of ethnic exogamy with all of the other Asian groups, suggesting a stronger boundary separating Filipinos from wider East/Southeast Asian panethnicity. This may reflect the unique Spanish cultural inheritance of the Philippines. To more formally test this hypothesis, I included a specific dummy variable for Latino/Filipino intermarriage and the result was a large and statistically significant difference from the overall odds of East/Southeast Asian/Latino intermarriage. The odds of Latino/Filipino intermarriage was 0.28 to 1, in comparison to an odds of about 0.13 to 1 for intermarriage between Latinos and other East/Southeast Asian groups.

The bottom panel of figure \@ref(fig:ethnic-heat-map) shows comparable results for the nine largest Latino ethnic groups. Overall, the odds ratios here are much lower than those for Asian ethnic groups reflecting the overall lower ethnic exogamy among Latino ethnic groups. The only case with no evidence of barriers to ethnic intermarriage was for Salvadorians and Guatemalans, where the estimated odds ratio is close to one.

The most notable feature of this figure is that ethnic affinities tend to cluster by region, with higher odds of intermarriage within the three groups of South American, Caribbean, and Mexican/Central American nationalities. The one exception to this pattern is for Cubans, who tend to be about equal distance from both the South American groups and other Caribbean groups. 

Figure \@ref(fig:racial-exogamy) shows the odds of intermarriage with Whites and Blacks for each of the East/Southeast Asian and Latino ethnic groups. for comparison, all of the ethnic exogamy odds ratios for a given ethnic group are also shown with small grey dots. 

```{r racial-exogamy, fig.cap="Odds of interracial marriage with Whites and Blacks for major Latino and E&SE Asian ethnic groups. For reference, odds of ethnic exogamy to other major ethnic groups are shown by small grey dots. Models are sorted from highest White/Black odds ratio to lowest. All models control for age differences, education differences, and language and birthplace endogamy."}
panethnic <- coefs_extended %>%
  filter(grepl(paste(c(latino_groups,asian_groups),collapse="|"), variable) & 
           !grepl("Black|White", variable) & 
           model=="+ both" & 
           data=="ACS 2014-18, expanded ethnic groups") %>%
  select(variable, coef)

#convert this to simple dataframe of all the coefs for each ethnic group
panethnic <- bind_rows(lapply(c(latino_groups,asian_groups), function(x) {
  return(tibble(ethnicity=x, 
                coef= panethnic$coef[str_detect(panethnic$variable, x)]))
}))
panethnic$group <- factor(panethnic$ethnicity %in% asian_groups,
                          levels=c(FALSE,TRUE), labels=c("Latino","E&SE ASian"))

intermar <- coefs_extended %>%
  filter(grepl(paste(c(latino_groups,asian_groups), collapse="|"), variable) & 
           grepl("Black|White", variable) & 
           model=="+ both" & 
           data=="ACS 2014-18, expanded ethnic groups") %>%
  mutate(ethnicity=str_remove(variable, "Black\\/|\\/White"),
         race=factor(str_detect(variable, "Black\\/"),
                     level=c(FALSE,TRUE), labels=c("White","Black")),
         group=factor(ethnicity %in% asian_groups,
                      levels=c(FALSE,TRUE), labels=c("Latino","E&SE ASian"))) %>%
  select(ethnicity, group, race, coef)

ggplot(intermar, 
       aes(y=reorder(ethnicity, coef, max), x=exp(coef)))+
  geom_vline(xintercept = 0, color="grey70", linetype=1)+
  geom_vline(xintercept = 1, color="grey70", linetype=3)+
  geom_point(aes(color=race), size=3)+
  geom_point(data=panethnic, alpha=0.5)+
  scale_color_manual(values=my_palette)+
  scale_x_continuous(limits=c(0,1.2))+
  facet_wrap(~group, scales = "free_y")+
  labs(y=NULL, 
       x="Odds ratio of interracial marriage relative to ethnic endogamy",
       color="Interracial partner")+
  theme_mine()+
  theme(legend.position = "bottom")
```

For all East/Southeast Asian and Latino ethnic groups, intermarriage with Whites is more likely than intermarriage with Blacks. The variation in the likelihood of interracial marriage across ethnic groups is more pronounced for Latino ethnic groups than for East/Southeast Asian groups. Among Latinos, Peruvians and Colombians are the most likely to intermarry with Whites with a odds ratios slightly higher than 0.6 to 1. At the other end of the spectrum the odds of Dominican/White intermarriage are slightly less than 0.25 to 1. For the five East/Southeast Asian groups, the odds ratios range from about 0.25 to 0.42.  

The odds of intermarriage with Blacks are uniformly low for all five East/Southeast Asian groups. The odds of intermarriage are also low for most Latino groups, with the exception of Dominicans and Puerto Ricans. For Dominicans, the odds of intermarriage with Blacks are almost identical to the odds of intermarriage with Whites. For Puerto Ricans, the odds of intermarriage with Blacks are the highest of any Latino ethnic group and only slightly less likely than the odds of intermarriage with Whites. These differences almost certainly reflect the much higher prevalence of Afrodescent in Puerto Rican and Dominican populations and the racialization of members of these populations as black within the US. However, this pattern is not observed for other Latino populations with a high prevalence of Afrodescent, such as Cubans and Colombians. 

The relative placement of ethnic exogamy odds ratios for each ethnic group in Figure \@ref(fig:racial-exogamy) is also telling. For all five Asian ethnic groups, every ethnic exogamy parameter is greater than the odds of intermarriage with Whites. For the Latino groups, these ethnic exogamy parameters are frequently in between the odds of White and Black intermarriage. For example, both Peruvians and Cubans are more likely to outmarry to a White person than to intermarry with any of the other Latino groups here. Mexicans are only slightly more likely to outmarry to most other Latino ethnic groups as they are to Whites, and are substantially less likely in two cases (Dominicans and Cubans). 

Overall, Figure \@ref(fig:racial-exogamy) implies very different patterns of intermarriage for Latino and East/Southeast Asian ethnic groups. For the East/Southeast Asian ethnic groups, we observe only small variation across groups and all groups follow the same general pattern. Intermarriage with all other ethnic groups is more likely than intermarriage with Whites or Blacks, and the barriers to intermarriage with Blacks are far more substantial than the barriers to intermarriage with Whites. The consistency of this pattern speaks to a broad panethnic pattern of intermarriage for East/Southeast Asians, even given some regional and national variation. 

On the other hand, the results for Latino groups are characterized by ethnic-specific heterogeneity. In no case do we see a clear separation in which ethnic exogamy is preferred to outmarriage with Whites. We also see large differences in the tendency to outmarry with Whites and Blacks. These results are not consistent with a broad tendency toward panethnicity among Latinos. Patterns of intermarriage instead seem to be very ethnically specific. 

# Conclusions

1.  Do we see evidence of pan-ethnicity in the marriage choices of Asians and Latinos and has this tendency changed over time?

2.  Answering this question is seriously complicated by the diversity of birthplace and language among and within Asian and Latino ethnic groups due to the continuing high levels of immigration from Asia and Latin America. The tendency for people to marry someone of the same birthplace and who speaks the same language is a significant barrier to pan-ethnic intermarriage for Asians, but not for Latinos.

3.  Once we account for birthplace and language endogamy, we see significant affinity between E/SE Asian and South Asian ethnic groups respectively. In the more recent ACS data, the odds of intermarrying with a member of a different ethnic group within the same pan-ethnic umbrella group were about 0.7 to 1 relative to marrying someone of the same exact ethnic group. In the case of E/SE Asians, where we can track this change over time, these odds have increased from about 0.5 to 1 in Census 1980. The results show little evidence of affinity between E/SE Asians and South Asians themselves, indicating that the growing pan-ethnicity among "Asians" is occurring in two distinct "melting pots."

4.  For Latinos, there is little evidence of change over time and the tendency toward ethnic exogamy is much lower than for Asians. Adjusting for birthplace and language endogamy, the odds of ethnic exogamy relative to ethnic endogamy for Latinos is roughly 0.25 to 1 in both time periods.

5.  It is important to recognize that these odds ratios reflect underlying affinities but not actual frequencies. In practice, the tendency toward ethnic exogamy is affected by birthplace and language diversity. This affect severely reduces the actual odds of ethnic exogamy for Asians much more than Latinos and produces odds ratios that are similar across all three groups, at about 0.25 to 1. Thus the higher underlying affinities among Asians would only be realized in a situation in which most Asians were US born and spoke English as a primary language, reflecting a situation of decreased immigration and greater acculturation.

6.  Given that we see no sign of declines in immigration, we are less likely to see clear pan-ethnic identities emerge from pan-ethnic intermarriage but rather what Massey has described as fractured identities among both Asians and Latinos.

7.  Why are the affinities among Latino ethnic groups lower than among Asian ethnic groups? This is somewhat surprising given that other research has noted the somewhat stronger basis for Latino pan-ethnicity. Language in particular has played a role in pan-ethnicity among Latinos, not simply in an ability to communicate in interpersonal situations, but in terms of creating pan-ethnic shared institutions. In some sense, language plays a role here in that it does not serve as a barrier to Latino groups in the same way that it does for Asians. However, the data clearly showed a weak affinity for ethnic exogamy among Latinos in comparison to Asians. This may reflect a deeper racialization of Asian identity. The idea of "Latino" as a racial category is still a contested idea and is contradicted by the phenotypic diversity of the Latino population, including but not limited to Afrodescent among some Latinos. On the other hand, "Asian" has a long history as a racial category in the US and more broadly, and was one of the original categories encoded by early racial taxonomic systems. In short, the experience of racialization may bind acculturated Asian Americans more closely together than ethnic identity. Latino identity, however, may remain more of a "federated" identity.

8.  Caveats about only one way to measure and identify pan-ethnicity.

# References
